---
title: "norming-spanish"
date: "2026-03-14"
output:
  pdf_document:
    toc: true
---

```{r packages, include = FALSE}
library(dplyr)
library(stringr)
library(tidyverse)
library(lme4)
library(magrittr)

theme_set(theme_bw())

```


## Exp info

- Experiment design 

The participants will sign a consent form, go through a headphone screening test, and then move on to the norming experiment. In this experiment,they  will go through 23 VOT steps x 4 repetitions/blocks x 3 word pairs + 12 catch trials  = 288 trials in total. Each participant hears all VOT steps across all word pairs (only within-subject). On catch trials, participants are instructed to press either left or right. 

The VOT steps are as follows: -100, -50, -20, -10, 0, 5, 10, 15, 20, 25, 30, 35, 40, 45, 50, 55, 60, 65, 70, 80, 90, 100, 120

- Randomization

The order of trials was randomized for each participant. Catch trials were distributed randomly throughout the experiment.

Participants were given the opportunity to take breaks after each block. Participants took about 20 minutes to complete all the trials. For each participant, /d/-initial words were either always displayed on the left side or always displayed on the right side. Across participants, this ordering was counter-balanced. 


## Preprocessing and exclusion

- Exclusion standards:

Participants who committed more than 3 errors out of the 12 catch trials will be excluded (N = 2).

Participants with an average RT more than three standard deviations from the mean of the by-participant means (N = 0) will be excluded.

Trials that are three SDs from the participant's mean RT were excluded from analysis (n=80) will be excluded.


```{r preprocess, warning=FALSE, include = FALSE}

raw <- read.csv("../norming-exp/data/data_exp_30subj_all_tasks.csv") %>%
  rename(ID = 'Participant.Private.ID') %>%
  rename(stim = 'Spreadsheet..stim') %>%
  rename(block = 'Spreadsheet..randomized_trial')%>%
  rename(rt = 'Reaction.Time')%>%
  filter(Display == "phonetic categorization") %>%
  filter(Response.Type == "response") %>%
  mutate(
    catch = case_when(
      str_detect(stim, "left")  ~ Spreadsheet..left,
      str_detect(stim, "right") ~ Spreadsheet..right,
    ))%>%
  mutate(catch.corr = if_else(catch == Response, 1, 0))%>%
  mutate(VOT = as.numeric(str_extract(stim, "(?<=_)[^.]+"))) %>%
  mutate(resp.num = ifelse(Response == "d", 0, 1))%>%
  mutate(word = str_extract(stim, "^[^_]+" )) 

```



```{r exclusion, warning=FALSE, include=FALSE}
# check acc on catch trials, exclude two participants
data.cal.acc <- raw %>% 
  filter(stim %in% c("pleft_ame.mp3", "pright_ame.mp3")) %>%
  group_by(ID) %>%
  summarise(exclude.catch = sum(catch.corr, na.rm = TRUE) < 10, .groups = "drop")
#summarise_at

data.exclude.catch <- data.cal.acc %>% 
  filter(exclude.catch == TRUE)%>% 
  distinct(ID)

# check participant-level rt,  all qualitfied
data.rt.part <- raw %>%
  left_join(data.cal.acc, by = "ID") %>% 
  filter(exclude.catch == FALSE) %>% 
  filter(stim != "pleft_ame.mp3" & stim != "pright_ame.mp3") %>%
  group_by(ID) %>% 
  mutate(rt.log = log10(ifelse(rt <= 0, NA, rt)),
    rt.log.z = as.numeric(scale(rt.log)),
    rt.log.z.mean = mean(rt.log.z, na.rm = T), 
    exclude.by.part.rt = ifelse(abs(scale(rt.log.z.mean)) > 3, TRUE, FALSE))

data.exclude.by.part.rt <- data.rt.part %>%
  filter(exclude.by.part.rt== TRUE)

# check trial-level rt, 80 trials excluded
data.rt.trial <- data.rt.part %>%
  mutate(exclude.by.trial.rt = ifelse(abs(rt.log.z) > 3, TRUE, FALSE))

data.exclude.by.trial.rt <- data.rt.trial %>% 
  filter(exclude.by.trial.rt == TRUE) 


# get clean data
data.preprc <- data.rt.trial %>% 
  filter(exclude.by.trial.rt == FALSE | exclude.by.part.rt == FALSE) %>%
  filter(stim != "pleft_ame.mp3" & stim != "pright_ame.mp3")  #exclude practice trials


```


## Plotting

General principles for wrangling and plotting

1. start from one data point

2. start from raw data

3. always include uncertainty


```{r wrangling}

# prepare the data for plot
data.plot <- data.preprc %>%
  select(VOT, resp.num, ID, block, word, stim, rt, rt.log)

summary(data.plot)
glimpse(data.plot)

```
### Start from single subject data

As we are concerned about the difference of categorization functions of three contrasts, I made "word" a grouping variable for every plot. 

Through data from the single subject, we can already see that the behavior of "din" contrast is different from that of "dip" and "dill": the number of 0, 1 response are closer to each other; (2) its overall mean is closer to 0.5, (3) its categorization curve shifts earlier  

```{r visualization single subj, fig.height=3, fig.width=4}

# choose 1 subject and summarise mean_resp
data.plot %>%
  filter(ID == '15432596')%>%
  group_by(word, VOT)%>%
  summarize(mean.resp.word = mean(resp.num),)

data.s1 <- data.plot %>%
  filter(ID == '15432596')

# simple count the number of 0, 1 responses for each word
p.s1.count <- ggplot(data.s1, aes(x = factor(resp.num), fill = word)) +
  geom_bar(position = "dodge") +
  labs(x = "response", y = "count")
p.s1.count # d is 0

# take a look at rt just in case
p.s1.rt <- ggplot(data.s1, aes(x = factor(word), y = rt.log)) +
  geom_point()+
  geom_boxplot(fill = NA, outlier.alpha = .1, notch = T)+
  labs(x = "word", y = "log rt")
p.s1.rt # d is 0.

# per subj, per word --> 69 data points each
p.s1.resp <- ggplot(data.s1, aes(x = factor(word), y = resp.num)) +
  stat_summary(fun.data = mean_cl_boot, geom = "pointrange", size = .2) +
  labs(x = "word", y = "mean") 
p.s1.resp

# per subj, per vot, per word -> 4 data points each
p.s1.vot <- ggplot(data.s1, aes(x = VOT, y = resp.num, color = word)) +
  stat_summary(fun.data = mean_cl_boot, geom = "pointrange", size = .2) +
  stat_summary(fun.data = mean_cl_boot, geom = "line") +
 # geom_smooth(method = "glm", method.args = list(family = "binomial")) #avoid making the plot too crowded
  
 #  # ver2, no uncertainty around mean
 #  stat_summary(fun = mean, geom = "point") +
 # stat_summary(fun = mean, geom = "line") +
 # geom_smooth(method = "glm", method.args = list(family = "binomial"))
  labs(x = "word", y = "mean") 
p.s1.vot


  
```


### move to multiple subj

Verify preliminary findings in multiple subjects.

```{r visualization multiple subj, fig.height=3, fig.width=4}

p.count = p.s1.count %+% 
  data.plot
p.count

p.rt = p.s1.rt %+% 
  data.plot
p.rt

p.resp = p.s1.resp %+% 
  data.plot
p.resp


p.vot = p.s1.vot %+% 
  data.plot
p.vot


```
Check subject-by-subject data. 

```{r each subj, fig.width=6, fig.height=16}
p.resp+
  facet_wrap(facets = ~ ID, ncol =3)
```








## Analysis

Our goal is to evaluate whether categorization functions for the three minimal pairs differs a lot from each other. And one minimal pair indeed differs. 


```{r logistic regression}
data.preprc %<>%
  mutate(VOT.z = as.numeric(scale(VOT)),
  word = factor(word))

model <- glmer(resp.num ~ VOT.z * word + (1 + VOT.z | ID),
  data = data.preprc,
  family = binomial)

summary(model)
```
