---
title: "tracker"
author: "Nan"
date: "2026-01-19"
output: html_document
---


```{r packages echo=FALSE}
library(dplyr)
library(stringr)
library(tidyverse)
library(lme4)
library(magrittr)
knitr::opts_chunk$set(warning = FALSE)

```





# create stimuli for VOT continuum

The process takes a voiced token and progressively deletes its vowel onset and replaces it with an approximately equivalent amount of onset taken from the word’s voiceless minimal-pair counterpart.

1.  Choose the word: bill/pill

2.  Annotate the repeitions of /b/, /p/ sounds to get a regression model for it

• Include negative VOT into the regression model

• Exclude pre-voicing from the first 25% of vowel duration

\todoNW{double check VOT is calculated separately}


```{r regression}


#df <- read.csv('../stimuli/recorded/speech-recording-b-p/resultsfile.txt')
df <- read.table( "../stimuli/recorded/speech-recording-b-p/resultsfile.txt", header = TRUE, sep = "\t")

#load vot
vot <- read.csv("../stimuli/recorded/speech-recording-b-p/vot-f0.csv") %>%
  select(files, VOT_ms)%>%
  #modify files to change from xx.wav to xx
  mutate(files = gsub(".wav", "", files))%>%
  #change colunmn name
  rename(Filename = files)


#merge vot into df
df <- merge(df, vot, by = "Filename")

# only include Filename starts with Bill or Pill
bpill <- df %>%
  filter(grepl("^bill|^pill", Filename))

bpill
```
> summary(bpill$f0_25._mean)
  Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
  98.07  105.78  107.21  108.60  112.26  118.63 

```{r regression model}

# run regression
model <- lm(f0_25._mean ~ VOT_ms, data = bpill)

# predict F0 at VOT = 130ms
predict(model, newdata = data.frame(VOT_ms = 130))
predict(model, newdata = data.frame(VOT_ms = 0))


```

result: 116.8766 for 130 ms VOT



3.  Select the base token to make the sound material:

• Speaker_Reinaldo

• Visual inspection (just for double check, which is typical in Spanish-accented English?): 1) All voiced sounds have prevoicing, 2) Voiceless sounds are less aspirated than English

• Evaluate naturalness: they all sound quite natural for me, but I chose the following two: 1) Bill_bk10 (shorter prevoicing), 2) Pill_bk8 (long VOT)

4.  Make the sound continuum for positive onset

\todoNW{i think we use cut-back here, use cut-back once again}


• cut back silent part from the sound

• For each minimal pair a continuum of 27 tokens was generated between 0ms and 130ms with a step-size of 5ms.

• Method of choosing aspiration: I chose from original words instead

• Min and max F0 onset: 109.22 – 113.87 (using absolute Hz)

Will's set up: dill/till (105.7Hz and 110.7 Hz), din/tin(107Hz and 112.2Hz), and dip/tip(103.6Hz and 108.9 Hz). The differences between the two endpoints are around 5-6Hz, which is similar to Tan & Jaeger's setup (246Hz and 252Hz)

\todoNW{faiil to predict at 0ms}

• F0 time range: 75 (default)

• F0 manipulation: relative to original F0

• VOT and F0 covariance: covary the two in tandem

• Vowel cut-back ratio: 0.33

• Aspiration intensity: 6 dB (the default; lowered by 6dB at Step 1 and gradually interpolated to increase till no attenuation at Step N); Attenuation of the aspiration continua in this experiment was made over 21 steps (0-100ms) which resulted in a range of 55.5 dB to 60.5 dB.

\todoNW{the start of burst; the end of aspiration, vowel duration excludes prevoicing}

5.  Make the sound continuum for negative onset

• Excise a sgement with clean-up prevoicing, note down its duration (-8.8ms)

• For 50ms pre-voicing from a 30ms sample: factor = 50/30 = 1.67, should be the duration factor (5.68 in my case)

• Take the corresponding time-scaled pre-voicing segment.

• Prepend it to the onset of the 0-ms VOT token.


\todoNW{Dr.VOT, check it up}
\todoNW{clean up blank before combining prevoicing to 0ms token}


6. rename files

```{r}
setwd('/Users/nwang40/Documents/Gitrepo/tan-jaeger/stimuli/d-t/dip_tip/!final VOT continuum_selected')
files <- list.files(pattern = "^dip_.*\\.wav$")

nums <- as.numeric(sub("^dip_(-?\\d+)\\.wav$", "\\1", files))
keep <- nums >= 0 & nums <= 27
files <- files[keep]
nums <- nums[keep]

new_files <- paste0("dip_+", nums * 5 - 5, ".wav")

data.frame(old = files, new = new_files)

file.rename(files, new_files)

```



# Norming experiment


```{r sound steps}


nums <- c(-100, -50, -20, -10, 0, 5, 10, 15, 20, 25, 30, 35, 40,
          45, 50, 55, 60, 65, 70, 80, 90, 100, 120)

files <- paste0("dill_", nums, ".wav")

```


- experiment design 

The participants will sign a consent form, go through a headphone screening test, and then move on to the norming experiment. In this experiment,they  will go through 23 VOT steps x 4 repetition/blocks x 3 word pair + 12 catch trials  = 288 trials in total. Each participant hears all VOT steps across all word pairs (only within-subject). On catch trials, participants are instructed to press either left or right. 

The VOT steps are as follows: -100, -50, -20, -10, 0, 5, 10, 15, 20, 25, 30, 35, 40, 45, 50, 55, 60, 65, 70, 80, 90, 100, 120

- randomization

The order of trials was randomized for each participant. Catch trials were distributed randomly throughout the experiment.

Participants were given the opportunity to take breaks after each block. Participants took about 20 minutes to complete all the trials. For each participant, /d/-initial words were either always displayed on the left side or always displayed on the right side. Across participants, this ordering was counter-balanced. 


```{r preprocess}

raw <- read.csv("../norming-exp/data/data_exp_30subj_all_tasks.csv") %>%
  rename(ID = 'Participant.Private.ID') %>%
  rename(stim = 'Spreadsheet..stim') %>%
  rename(block = 'Spreadsheet..randomized_trial')%>%
  rename(rt = 'Reaction.Time')%>%
  filter(Display == "phonetic categorization") %>%
  filter(Response.Type == "response") %>%
  mutate(
    catch = case_when(
      str_detect(stim, "left")  ~ Spreadsheet..left,
      str_detect(stim, "right") ~ Spreadsheet..right,
    ))%>%
  mutate(catch.corr = if_else(catch == Response, 1, 0))%>%
  mutate(VOT = as.numeric(str_extract(stim, "(?<=_)[^.]+"))) %>%
  mutate(resp.num = ifelse(Response == "d", 0, 1))%>%
  mutate(word = str_extract(stim, "^[^_]+" )) 

```

- Exclusions

Participants who committed more than 3 errors out of the 12 catch trials will be excluded (N = 2).

Participants with an average RT more than three standard deviations from the mean of the by-participant means (N = 0) will be excluded.

Trials that are three SDs from the participant's mean RT were excluded from analysis (N=) will be excluded

```{r catch trials}
# check acc on catch trials, exclude two participants
data.cal.acc <- raw %>% 
  filter(stim %in% c("pleft_ame.mp3", "pright_ame.mp3")) %>%
  group_by(ID) %>%
  summarise(exclude.catch = sum(catch.corr, na.rm = TRUE) < 10, .groups = "drop")

data.exclude.catch <- data.cal.acc %>% 
  filter(exclude.catch == TRUE)%>% 
  distinct(ID)

# check participant-level rt,  all qualitfied
data.rt.part <- raw %>%
  left_join(data.cal.acc, by = "ID") %>% 
  filter(exclude.catch == FALSE) %>% 
  filter(stim != "pleft_ame.mp3" & stim != "pright_ame.mp3") %>%
  group_by(ID) %>% 
  mutate(rt.log = log10(ifelse(rt <= 0, NA, rt)),
    rt.log.z = as.numeric(scale(rt.log)),
    rt.log.z.mean = mean(rt.log.z, na.rm = T), 
    exclude.by.part.rt = ifelse(abs(scale(rt.log.z.mean)) > 3, TRUE, FALSE))

data.exclude.by.part.rt <- data.rt.part %>%
  filter(exclude.by.part.rt== TRUE)

# check trial-level rt, 80 trials excluded
data.rt.trial <- data.rt.part %>%
  mutate(exclude.by.trial.rt = ifelse(abs(rt.log.z) > 3, TRUE, FALSE))

data.exclude.by.trial.rt <- data.rt.trial %>% 
  filter(exclude.by.trial.rt == TRUE) 


# get clean data
data.preprc <- data.rt.trial %>% 
  filter(exclude.by.trial.rt == FALSE | exclude.by.part.rt == FALSE) %>%
  filter(stim != "pleft_ame.mp3" & stim != "pright_ame.mp3")  #exclude practice trials

```



```{r visualization}

# prepare the data for plot
data.plot <- data.preprc %>%
  select(VOT, resp.num, ID, block, word, stim)


# Plot the raw data for visual inspection
p.raw <- ggplot(data.plot, aes(x = factor(ID), y = resp.num)) +
  stat_summary(fun = mean, geom = "line") +
  stat_summary(fun = mean, geom = "point") +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 1)+
  labs(x = "ID", y = "mean")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p.raw

# plot group-level data
p.group <- ggplot(data.plot, aes(x = VOT, y = resp.num)) +
  stat_summary(fun = mean, geom = "line") +
  stat_summary(fun = mean, geom = "point") +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 1)+
  geom_smooth()+
  labs(x = "VOT", y = "probability of /t/")

p.group

# group by each continua
p.group + facet_wrap(~ word) 

p.group + facet_wrap(~ block) 

```
Our goal is to evaluate whether categorization functions for the three minimal pairs differs a lot from each other. 

```{r logistic regression}


data.preprc %<>%
  mutate(VOT.z = as.numeric(scale(VOT)),
  word = factor(word))

model <- glmer(resp.num ~ VOT.z * word + (1 + VOT.z | ID),
  data = data.preprc,
  family = binomial)

summary(model)
```








## Stimulus inventory

Tan stimuli: 

- 4 word pair (dill–till, 1din–tin, dip–tip); VOT continua from −100 ms to +130 ms (5ms step size)

- 3 exposure blocks (54 trials each); 6 test blocks (12 trials each) = 234 trials in general

- In each exposure block: (1) 24/d/ and 24/t/ trials forming distributions over VOT, as well as 6 catch trials; (2) half of the /d/ and half of the /t/ trials in each exposure block were labeled, and the other half were unlabeled.

- In each test block: (1) trials are identical within and across conditions, but randomized; (2) 12 VOT values: −5, 5, 15, 25, 30, 35, 40, 45, 50, 55, 65, 70 ms

- 36 exposure-test combinations: 4 tests x 3 exposure conditions x 3 Latin square-designed block orders


Our stimuli: 

- 30 VOT x 3 word pair (dill/till, din/tin, dip/tip) 


# Experiment design

## Trial structure

Each trial started with a dark-shaded green fixation dot being displayed. 

At 500 ms from trial onset, two minimal pair words appeared on the screen. 

At 1000 ms from trial onset, the fixation dot would turn bright green and participants had to click on the dot to play the recording.(This was meant to reduce trial-to-trial correlations by resetting the mouse pointer to the center of the screen at the start of each trial.)


## Experiment structure

Pass a headphone test in order to continue (Woods, Siegel, Traer, & McDermott, 2017).

Take breaks every 60 trials.

Completed an exit survey and an optional demographics survey.


# Counterbalancing and randomization

Within each test block: fixed, not randomized across participants;

Within each exposure block: three VOT assignment randomized across participants.



#

