---
title: "chodroff2017"
author: "Nan"
format:
  pdf:
    toc: true
    toc-depth: 4
    number-sections: true
    include-in-header:
      text: |
        \usepackage{underscore}
        \newcommand{\todo}[1]{\textbf{\color{red}[TODO: #1]}}
        \newcommand{\todoflorian}[1]{\textbf{\color{blue}[TODO florian: #1]}}
fontsize: 10pt
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(ggplot2)
library(Hmisc)     # mean_cl_boot, for bootstrapped confidence intervals
library(knitr)     # kable, for tables in the rendered report
library(broom)     # tidy(), to collect t-test results into a data frame

# format a p-value APA-style for inline reporting
format_p <- function(p.value) {
  if (p.value < .001) "< .001" else paste0("= ", formatC(p.value, format = "f", digits = 3))
}

# compact "t(df) = x.xx, p ..." string from one tidy t-test row, for inline use
report_row <- function(tidy.row) {
  sprintf("*t*(%.1f) = %.2f, *p* %s",
          tidy.row$parameter, tidy.row$statistic, format_p(tidy.row$p.value))
}

# unit label for each measure, used in the table and the figure strips
measure.unit <- c("VOT" = "VOT (ms)",
                  "Vowel duration" = "Vowel duration (ms)",
                  "F0" = "F0 (Hz)")
```

# Dataset overview

The dataset accompanies Chodroff & Wilson's work on the structure of laryngeal contrast and voice-onset time in American English. Twenty-eight native English speakers produced monosyllabic words beginning with one of the six English stops (/p t k b d g/). For every stop token the recordings were force-aligned and a Praat script measured the stop interval, the duration of the following sonorant, the containing word, and a ten-point fundamental frequency contour sampled every 5 ms from vowel onset.

## Measures analysed

We contrast three acoustic measures across talker gender:

- **VOT** (ms) — the stop's voice-onset time, i.e. the interval from burst release to voicing onset, the standard voicing cue for stops. In this corpus the annotated stop interval *is* the VOT, so VOT and the stop's own segment duration are the same quantity here.
- **Vowel duration** (ms) — the length of the vowel that follows the stop, taken from the following-sonorant duration. As the README notes, once the consonant-cluster ("L") tokens are removed the following sonorant is always the vowel.
- **F0** (Hz) — the fundamental frequency at vowel onset, summarised as the mean of the first five 5-ms samples (the first 25 ms of the vowel), following the corpus convention. F0 largely reflects vocal-fold anatomy, so it is included partly as a reference measure (see the interpretation).

VOT is a property of the consonant's laryngeal timing, vowel duration reflects the length of the following vowel, and F0 reflects vocal-fold vibration rate. Comparing all three shows whether a gender difference lives in the consonant cue, the vowel, the pitch, or some combination.

## Preprocessing

The raw data is preprocessed following the steps below:

- Select the target stops, i.e. /t/, /d/;
- Remove consonant-cluster tokens. Cluster words such as *KLOT* or *PLOT* (CLVC) are flagged by a following sonorant of `"L"`; these are dropped so that the segment following every retained stop is a vowel and all three measures are comparable.
- VOT and the following-vowel duration are stored in seconds and re-expressed in milliseconds.
- The ten F0 samples are stored as text with `"--undefined--"` where Praat could not track pitch. These frames are read as missing, as are physiologically implausible values (outside 50–400 Hz, which are octave-tracking artefacts). F0 is then summarised as the mean of the first five samples (first 25 ms of the vowel); the few tokens left with no usable F0 sample are dropped from the F0 comparison only.


```{r load-preprocessing}

data.dir     <- "../data/chodroff2017"
cue.file     <- file.path(data.dir, "cueAnalysis_engCVC.txt")
speaker.file <- file.path(data.dir, "engCVC_speakerList.csv")

column.names <- c("file", "stop", "start", "end", "interval", "vot",
                  "following.sonorant", "sonorant.dur", "word", "word.dur",
                  paste0("f0_", 1:10))

engcvc.raw <- read_tsv(cue.file, col_names = column.names, show_col_types = FALSE)

# keep /t/ and /d/, drop CLVC (following "L"), derive speaker; measures in their units
td.before.cluster.removal <- engcvc.raw %>%
  filter(stop %in% c("T", "D"))

td.tokens <- td.before.cluster.removal %>%
  filter(following.sonorant != "L") %>%
  # F0 columns are text with "--undefined--". Make them numeric, and set both
  # undefined frames and physiologically implausible values (outside 50-400 Hz,
  # i.e. Praat octave-tracking errors) to NA so they drop out of the average.
  mutate(across(f0_1:f0_5, ~ {
    hz <- suppressWarnings(as.numeric(na_if(.x, "--undefined--")))
    ifelse(!is.na(hz) & hz >= 50 & hz <= 400, hz, NA_real_)
  })) %>%
  mutate(subj         = str_remove(file, "_edited"),
         vot.ms       = vot * 1000,                              # consonant voicing cue
         vowel.dur.ms = sonorant.dur * 1000,                     # following-vowel duration
         f0.mean.hz   = rowMeans(across(f0_1:f0_5), na.rm = TRUE),# mean F0 over first 25 ms
         stop         = factor(stop, levels = c("T", "D")))


# attach gender, keep known-gender speakers only
speaker.gender <- read_csv(speaker.file, show_col_types = FALSE)

td.tokens.gendered <- td.tokens %>%
  left_join(speaker.gender, by = "subj") %>%
  filter(!is.na(gender)) %>%
  mutate(gender = factor(gender, levels = c("female", "male")))


# reshape to long: one row per token x measure, so the pipeline below is written
# once and carries the extra `measure` grouping. rowMeans of an all-undefined
# token gives NaN (caught by is.na), so a value-drop leaves F0-less tokens out
# of the F0 rows only.
measures.long <- td.tokens.gendered %>%
  transmute(subj, gender, stop,
            VOT              = vot.ms,
            `Vowel duration` = vowel.dur.ms,
            F0               = f0.mean.hz) %>%
  pivot_longer(cols = c(VOT, `Vowel duration`, F0),
               names_to = "measure", values_to = "value") %>%
  filter(!is.na(value)) %>%
  mutate(measure = factor(measure, levels = c("VOT", "Vowel duration", "F0")))


# Bookkeeping counts, for the auditable prose that follows
n.tokens.raw            <- nrow(engcvc.raw)
n.cluster.removed       <- sum(td.before.cluster.removal$following.sonorant == "L")
n.unknown.gender.tokens <- nrow(td.tokens) - nrow(td.tokens.gendered)
n.td.tokens             <- nrow(td.tokens.gendered)
n.td.speakers           <- n_distinct(td.tokens.gendered$subj)
n.f0.dropped            <- sum(is.nan(td.tokens.gendered$f0.mean.hz))

# Sanity checks — durations positive, F0 in a plausible human range, all labelled
stopifnot(all(td.tokens.gendered$vot.ms       > 0),
          all(td.tokens.gendered$vowel.dur.ms > 0),
          all(filter(measures.long, measure == "F0")$value %>% between(50, 400)),
          all(!is.na(measures.long$gender)))

```

From the `r n.tokens.raw` raw stop tokens, restricting to /t/ and /d/ and dropping
`r n.cluster.removed` consonant-cluster tokens leaves `r n.td.tokens` tokens from
`r n.td.speakers` talkers (`r n.f0.dropped` of which lack a usable F0 sample and
are excluded from the F0 comparison only). Their distribution across stop and
talker gender is shown in Table \ref{tbl-counts}.

```{r}
#| label: tbl-counts
#| tbl-cap: "Number of /t/ and /d/ tokens and contributing talkers, by talker gender, in the final analysis set."
counts.by.cell <- td.tokens.gendered %>%
  group_by(stop, gender) %>%
  summarise(n.tokens   = n(),
            n.speakers = n_distinct(subj),
            .groups = "drop")

kable(counts.by.cell,
      col.names = c("Stop", "Gender", "Tokens", "Talkers"),
      align = "llrr")
```

# Analyze phoneme duration, VOT, and F0

## Goal and motivation

The goal is to compare three acoustic measures — the stop's **VOT**, the following
**vowel duration**, and the vowel-onset **F0** — between female and male talkers,
separately for /t/ and /d/. VOT indexes the laryngeal timing of the consonant,
vowel duration indexes the length of the following vowel, and F0 indexes the rate
of vocal-fold vibration. Examining all three lets us see which properties carry a
gender difference. F0 is expected to differ substantially for anatomical reasons
and so serves as much as a reference (a "does the gender split behave as known?"
check) as a finding in its own right.

The comparison is made on **per-talker means rather than raw tokens**. Each talker
contributes many tokens, and tokens from the same talker are not independent;
collapsing to one mean per talker (for each stop and measure) yields a single
independent observation per talker, the appropriate unit for a between-groups test.
We then compare the female and male talker means with a two-sided **Welch
two-sample t-test** (unequal-variance form), run once for each of the six
stop × measure combinations.

## Results

```{r analysis}
# ---------------------------------------------------------------------------
# Speaker-level means: one mean per talker x stop x measure = the analysis unit
# ---------------------------------------------------------------------------
speaker.means <- measures.long %>%
  group_by(subj, gender, stop, measure) %>%
  summarise(mean.value = mean(value),
            n.tokens   = n(),
            .groups = "drop")

# ---------------------------------------------------------------------------
# Group-level summary: mean of the talker means, and SD across talkers
# ---------------------------------------------------------------------------
group.summary <- speaker.means %>%
  group_by(measure, stop, gender) %>%
  summarise(n.speakers        = n(),
            mean.value        = mean(mean.value),
            sd.across.talkers = sd(mean.value),
            .groups = "drop")

# ---------------------------------------------------------------------------
# Welch t-test of female vs. male talker means, once per stop x measure.
# broom::tidy() collects all six tests into one data frame. In each row
# estimate1 = female mean, estimate2 = male mean, estimate = female - male.
# ---------------------------------------------------------------------------
test.results <- speaker.means %>%
  group_by(measure, stop) %>%
  group_modify(~ tidy(t.test(mean.value ~ gender, data = .x))) %>%
  ungroup()

# individual rows pulled out for inline reporting
vot.t.row   <- filter(test.results, measure == "VOT",            stop == "T")
vot.d.row   <- filter(test.results, measure == "VOT",            stop == "D")
vowel.t.row <- filter(test.results, measure == "Vowel duration", stop == "T")
vowel.d.row <- filter(test.results, measure == "Vowel duration", stop == "D")
f0.t.row    <- filter(test.results, measure == "F0",             stop == "T")
f0.d.row    <- filter(test.results, measure == "F0",             stop == "D")
```

```{r}
#| label: tbl-results
#| tbl-cap: "Female vs. male comparison of VOT, vowel duration, and F0 for /t/ and /d/, on per-talker means. Columns give the female and male group means, their difference (female - male), the Welch t statistic, degrees of freedom, p-value, and the 95% CI on the difference. Units follow the Measure column: ms for VOT and vowel duration, Hz for F0."
test.results %>%
  transmute(Measure     = measure.unit[as.character(measure)],
            Stop        = recode(stop, T = "/t/", D = "/d/"),
            Female      = round(estimate1, 1),
            Male        = round(estimate2, 1),
            Diff        = round(estimate, 1),
            t           = round(statistic, 2),
            df          = round(parameter, 1),
            p           = ifelse(p.value < .001, "< .001",
                                 formatC(p.value, format = "f", digits = 3)),
            `95% CI`    = sprintf("[%.1f, %.1f]", conf.low, conf.high)) %>%
  arrange(Measure, Stop) %>%
  kable(align = "llrrrrrrl")
```

The three measures behave differently. **VOT** shows a significant female-longer
difference for both stops: /t/ `r round(vot.t.row$estimate1, 1)` vs.
`r round(vot.t.row$estimate2, 1)` ms (`r report_row(vot.t.row)`); /d/
`r round(vot.d.row$estimate1, 1)` vs. `r round(vot.d.row$estimate2, 1)` ms
(`r report_row(vot.d.row)`). **Vowel duration** shows *no* reliable gender
difference for either stop (/t/ `r report_row(vowel.t.row)`; /d/
`r report_row(vowel.d.row)`), with confidence intervals spanning zero. **F0**, as
expected, is much higher for female talkers — /t/ `r round(f0.t.row$estimate1, 0)`
vs. `r round(f0.t.row$estimate2, 0)` Hz (`r report_row(f0.t.row)`); /d/
`r round(f0.d.row$estimate1, 0)` vs. `r round(f0.d.row$estimate2, 0)` Hz
(`r report_row(f0.d.row)`) — a difference of roughly
`r round(f0.t.row$estimate1 / f0.t.row$estimate2, 2)`× in pitch. The full set of
estimates and tests is in Table \ref{tbl-results}, and the underlying per-talker
distributions are shown in Figure \ref{fig-all}.

```{r}
#| label: fig-all
#| fig-cap: "VOT (top, ms), following-vowel duration (middle, ms), and vowel-onset F0 (bottom, Hz) by talker gender, for /t/ (left) and /d/ (right). Each coloured point is one talker's mean; the black point and bars are the group mean and a bootstrapped 95% confidence interval across talkers (mean_cl_boot). Every panel has its own y-axis, and the unit differs by row (see the strip labels). Female and male talkers separate clearly on VOT and F0, but overlap on vowel duration."
#| fig-width: 8
#| fig-height: 9.5

# palette defined once so gender reads the same colour in every figure
gender.palette <- c(female = "#C2185B", male = "#1565C0")

ggplot(speaker.means, aes(x = gender, y = mean.value, colour = gender)) +
  # each dot = one talker's mean for this stop x measure
  geom_jitter(width = 0.12, height = 0, alpha = 0.5, size = 2) +
  # group mean + bootstrapped 95% CI across talkers
  stat_summary(fun.data = mean_cl_boot, geom = "errorbar",
               width = 0.15, linewidth = 0.8, colour = "black") +
  stat_summary(fun = mean, geom = "point", size = 3, colour = "black") +
  facet_wrap(~ measure + stop, scales = "free_y", nrow = 3,
             labeller = labeller(measure = as_labeller(measure.unit),
                                 stop    = as_labeller(c(T = "/t/", D = "/d/")))) +
  scale_colour_manual(values = gender.palette) +
  labs(x = "Talker gender", y = "Talker mean (unit varies by row)", colour = "Gender") +
  theme_bw(13) +
  theme(legend.position = "none")
```

## Interpretation

Bringing the three measures together sharpens the picture. The gender difference
is **carried by the consonant's VOT and by F0, but not by vowel duration**. Female
talkers have reliably longer VOT for both stops (about 20 ms for the aspirated
/t/, smaller in absolute terms for /d/), and much higher F0 — roughly
`r round(f0.t.row$estimate1 / f0.t.row$estimate2, 2)`× that of male talkers — yet
their following vowels are no longer than male talkers'.

The F0 and VOT differences are not on the same footing, however. The F0 gap is
very large and physiologically expected: adult female talkers have shorter, less
massive vocal folds that vibrate faster, so a higher F0 is essentially guaranteed
and functions here as a sanity check that the gender labelling behaves as known,
rather than as a discovery about /t/ and /d/. The VOT difference is the more
substantive result: it concerns the *timing* of the consonant rather than talker
anatomy, and the fact that vowel duration does *not* differ argues against a simple
"female talkers lengthen everything" account — if that were the mechanism, the
vowels would lengthen too.

Two caveats temper the result: the tests rest on `r n.td.speakers` talkers, so the
confidence intervals are correspondingly wide; and because talker gender is
confounded with any anatomical or speaking-rate differences that covary with it in
this sample, the analysis establishes an association rather than a cause. Within
those limits, talker gender is a meaningful source of variation in the VOT (and,
trivially, the F0) of /t/ and /d/ in the `engCVC` corpus, but not in the duration
of the following vowel.

# Correlation between VOT and vowel duration

## Goal and motivation

Having compared the measures separately across gender, we now ask whether two of
them covary *within* a token: do stop tokens with longer VOT also have longer
following vowels? Two design choices shape the analysis. First, the correlation is
computed **within each stop**. /t/ has both a long VOT and its own vowel-duration
distribution, so pooling /t/ and /d/ would create a between-category correlation
that says nothing about the within-category relationship — a Simpson's-paradox
trap. Second, we examine the relationship at two levels: the **token level** (every
stop token) and the **talker-mean level** (one point per talker), since "are
long-VOT *tokens* long-vowel tokens?" and "do long-VOT *talkers* have long vowels?"
are different questions.

## Token-level relationship

```{r}
#| label: fig-cor-token
#| fig-cap: "Token-level relationship between VOT (x) and following-vowel duration (y), for /t/ (left) and /d/ (right). Each point is one stop token; the line is an OLS fit with its 95% confidence band, and the annotation gives the within-stop Pearson correlation. The x-axis is free between panels because /t/ and /d/ occupy different VOT ranges. Both correlations are essentially zero."
#| fig-width: 8
#| fig-height: 4

# Pearson r of VOT vs. vowel duration within each stop (reused inline below)
cor.by.stop.token <- td.tokens.gendered %>%
  group_by(stop) %>%
  summarise(r = cor(vot.ms, vowel.dur.ms),
            p = cor.test(vot.ms, vowel.dur.ms)$p.value,
            x = min(vot.ms), y = max(vowel.dur.ms),   # anchor annotation to panel
            .groups = "drop") %>%
  mutate(label = sprintf("r = %.2f, p %s", r,
                         ifelse(p < .001, "< .001",
                                paste0("= ", formatC(p, format = "f", digits = 3)))))

cor.tok.t <- filter(cor.by.stop.token, stop == "T")
cor.tok.d <- filter(cor.by.stop.token, stop == "D")

ggplot(td.tokens.gendered, aes(x = vot.ms, y = vowel.dur.ms)) +
  geom_point(alpha = 0.25, size = 1) +
  geom_smooth(method = "lm", se = TRUE, colour = "grey20") +
  geom_text(data = cor.by.stop.token, aes(x = x, y = y, label = label),
            hjust = 0, vjust = 1, size = 4, fontface = "italic") +
  facet_wrap(~ stop, scales = "free_x",
             labeller = as_labeller(c(T = "/t/", D = "/d/"))) +
  labs(x = "VOT (ms)", y = "Following-vowel duration (ms)") +
  theme_bw(13)
```

At the token level the two durations are essentially unrelated within each stop:
the correlation is `r sprintf("%.2f", cor.tok.t$r)` for /t/ (*p* `r format_p(cor.tok.t$p)`)
and `r sprintf("%.2f", cor.tok.d$r)` for /d/ (*p* `r format_p(cor.tok.d$p)`). These are
distinguishable from zero only because each stop contributes about a thousand
tokens; they explain on the order of 1–3% of the variance, so there is no
practically meaningful token-level coupling between VOT and the length of the
following vowel.

## Talker-level relationship

```{r}
#| label: fig-cor-speaker
#| fig-cap: "Talker-level relationship between mean VOT (x) and mean following-vowel duration (y): one point per talker, for /t/ (left) and /d/ (right), with an OLS fit, its 95% band, and the within-stop Pearson correlation across talkers. Free x-axis as above. Neither correlation is reliable."
#| fig-width: 8
#| fig-height: 4

# one point per talker x stop
speaker.xy <- td.tokens.gendered %>%
  group_by(subj, stop) %>%
  summarise(vot = mean(vot.ms), vowel = mean(vowel.dur.ms), .groups = "drop")

cor.by.stop.speaker <- speaker.xy %>%
  group_by(stop) %>%
  summarise(r = cor(vot, vowel),
            p = cor.test(vot, vowel)$p.value,
            x = min(vot), y = max(vowel), .groups = "drop") %>%
  mutate(label = sprintf("r = %.2f, p %s", r,
                         ifelse(p < .001, "< .001",
                                paste0("= ", formatC(p, format = "f", digits = 3)))))

cor.spk.t <- filter(cor.by.stop.speaker, stop == "T")
cor.spk.d <- filter(cor.by.stop.speaker, stop == "D")

ggplot(speaker.xy, aes(x = vot, y = vowel)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = TRUE, colour = "grey20") +
  geom_text(data = cor.by.stop.speaker, aes(x = x, y = y, label = label),
            hjust = 0, vjust = 1, size = 4, fontface = "italic") +
  facet_wrap(~ stop, scales = "free_x",
             labeller = as_labeller(c(T = "/t/", D = "/d/"))) +
  labs(x = "Talker mean VOT (ms)", y = "Talker mean vowel duration (ms)") +
  theme_bw(13)
```

The talker-level picture agrees: across the `r n.td.speakers` talkers the
correlation between mean VOT and mean vowel duration is `r sprintf("%.2f", cor.spk.t$r)`
for /t/ (*p* `r format_p(cor.spk.t$p)`) and `r sprintf("%.2f", cor.spk.d$r)` for /d/
(*p* `r format_p(cor.spk.d$p)`), neither reliably different from zero. So talkers
with longer VOT are not systematically the ones with longer vowels — consistent
with the earlier finding that the gender effect lives in VOT specifically and does
not carry over to vowel duration.

