---
title: "formal-spanish"
format: html
editor: visual
---

```{r load library and functions from TJ OSF, include=FALSE}
rm(list = ls())
library(dplyr)
library(stringr)
library(tidyverse)
library(lme4)
library(magrittr)
library(ggeffects)
library(MASS)
library(emmeans)
library(brms)
library(tidybayes)
library(patchwork)
source("functions.R")
```

```{r preamble setup, include=FALSE}

chains <- 4
options(
  width = 1000,
  mc.cores = min(chains, parallel::detectCores())
)

GLOBAL_SEED <- 920
set.seed(42007)

myGplot.defaults("paper")
colours.condition <- c("expo-20ms" = "#D32F2F", "expo-10ms" = "#2E7D32", "expo+20ms" = "#1E88E5")

PREAMBLE_LOADED <- TRUE

```

```{r load data (old), include=FALSE}

#currently disable as we'll directly use the preprocessed dataset, but keep this for record

if (FALSE) {
raw <- read.csv("../data/formal/formal-full-data-preprocessed.csv") %>%
  rename(ID = 'Participant.Private.ID',
         list = 'counterbalance.gj4p',
         word = 'Spreadsheet..word',
         stim = 'Spreadsheet..stim',
         trial_type = 'Spreadsheet..trial_type',
         blockID = 'Spreadsheet..blockID',
         rt = 'Reaction.Time', 
         condition = 'Spreadsheet..condition',
         VOT = 'Spreadsheet..VOT')%>%
  filter(Display == "phonetic categorization") %>%
  filter(Screen == "2AFC") %>%
  filter(Response.Type == "response") %>%
  #filter(Response != "") %>%                    # drop screen-advance events (blank Response)
  #mutate(
  #  catch = case_when(
  #    str_detect(stim, "left")  ~ Spreadsheet..left,
  #    str_detect(stim, "right") ~ Spreadsheet..right,
  #  ))%>%
  #mutate(catch.corr = if_else(catch == Response, 1, 0))%>%
  #mutate(VOT = as.numeric(str_extract(stim, "(?<=_)[^.]+"))) %>%
  mutate(resp.num = case_when(
      str_detect(Response, "^t") ~ 1,
      str_detect(Response, "^d") ~ 0))
  #mutate(word = str_extract(stim, "^[^_]+" ))

participant.condition <- raw %>%
  filter(condition != "") %>%
  distinct(ID, condition) %>%
  rename(participant.condition = condition) # i move it upwards, so no need for left_join every time @will 

raw <- raw %>% left_join(participant.condition, by = "ID") 
}
```

```{r save pre-processed dataset (need to be <100MB to be pushed to github), include=FALSE}

#already run, no need to implement
#write.csv(raw, "../data/formal/formal-full-data-preprocessed.csv", row.names = FALSE)

```

```{r load preprocessed dataset}
raw <- read.csv("../data/formal/formal-full-data-preprocessed.csv")
raw$rt.log = log(raw$rt)
raw <- raw |> 
  rename(label_type = 'Spreadsheet..label')
```

```{r sanity check of experiment setup}

#check counterbalanced list assignment
sum.list.assign <- raw %>%
  group_by(list) %>%
  summarise(n_subjects = n_distinct(ID))

#check exposure condition assignment
sum.condition.assign <- raw %>%
  filter(condition!="") %>%
  group_by(condition) %>%
  summarise(n_subjects = n_distinct(ID))

#check trial number in each exposure and test blocks
sum.trial.num <- raw %>%
  filter(trial_type != "catch") %>% 
  group_by(ID, blockID) %>% 
  summarise(
    trial_num = n(),
    .groups = "drop"
  ) %>% 
  tidyr::pivot_wider(
    names_from = blockID, 
    values_from = trial_num
  )

```

```{r participant-level check and exclusion}

# (1) check catch trial accuracy
sum.catch_accuracy <- raw %>%
  filter(trial_type == "catch") %>% 
  mutate(catch.corr = if_else(word == Response, 1, 0)) %>% 
  group_by(ID) %>% 
  summarise(
    total_catch = n(),
    accuracy = mean(catch.corr),
    exclude.catch  = accuracy < 0.83
  )

data.exclude.catch.low.acc <- sum.catch_accuracy %>% 
  filter(exclude.catch == TRUE)%>% 
  distinct(ID)


# (2) check labelled trial accuracy
sum.labelled_accuracy <- raw %>%
  filter(label_type == "labeled") %>% 
  mutate(label.corr = if_else(word == Response, 1, 0)) %>% 
  group_by(ID) %>% 
  summarise(
    total_labelled = n(),
    accuracy = mean(label.corr),
    exclude.labelled  = accuracy < 0.94
  )

data.exclude.labelled.low.acc <- sum.labelled_accuracy %>% 
  filter(exclude.labelled == TRUE)%>% 
  distinct(ID)


# (3a) check valid responses across participants
sum.resp.num <- raw %>%
  filter(trial_type != "catch", !is.na(resp.num)) %>%
  group_by(ID) %>% 
  summarise(
    resp.num = sum(!is.na(resp.num)),
    .groups = "drop"
  )

# (3b) check NA responses across participants
sum.na.resp <- raw %>% 
  filter(trial_type != "catch", is.na(resp.num)) %>% 
  count(ID, name = "na_count") %>%
  mutate(
    exclude.high.na = na_count > 21
  )

data.exclude.high.na <- sum.na.resp %>% 
  filter(exclude.high.na == TRUE)%>% 
  distinct(ID)


# (4) check overall RT across participants
sum.subj.RT <- raw %>%
  filter(trial_type != "catch") %>% 
  group_by(ID) %>% 
  summarise(
    trial_num = n(),
    subj_mean_RT = mean(rt, na.rm = TRUE),
    subj_mean_RT.log = mean(rt.log, na.rm = TRUE),
    .groups = "drop"
  ) |> 
  mutate(
    subj_mean_RT.log.z = as.numeric(scale(subj_mean_RT.log)),
    exclude.by.subj.rt = abs(subj_mean_RT.log.z) > 3
  )

data.exclude.by.subj.rt <- sum.subj.RT |> 
  filter(exclude.by.subj.rt == TRUE)%>% 
  distinct(ID)


# (5) check atypical categorization responses based on T1 block
# note that each VOT token appeared in a test block only once
sum.cat.diff <- raw %>%
  filter(trial_type != "catch", blockID == "T5") %>%
  mutate(
    vot_group = case_when(
      VOT <= 30 ~ "short_VOT", #bin responses for VOT= −5, 5, 15, 25, 30 together
      VOT >= 45 ~ "long_VOT",  #bin responses for VOT= 45, 50, 55, 65, 70 together
      TRUE ~ "middle_VOT"      #set aside VOT= 35, 40 which might be in the ambiguous region
    )
  ) %>%
  filter(vot_group != "middle_VOT") %>%
  group_by(ID, vot_group) %>%
  summarise(mean_resp = mean(resp.num, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = vot_group, values_from = mean_resp) %>%
  mutate(
    category_diff = long_VOT - short_VOT,
    exclude.atypical.cat = category_diff <= 0.3
  )

flagged.atypical.cat.ids <- sum.cat.diff |> 
  filter(exclude.atypical.cat == TRUE) %>% 
  distinct(ID) %>%
  pull(ID)

# check performance in the T1 block for flagged ids
raw %>%
  filter(trial_type != "catch", blockID == "T5") %>%
  filter(ID %in% flagged.atypical.cat.ids) %>%
  ggplot(aes(x = VOT, y = resp.num)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", se = FALSE, color = "firebrick") +
  facet_wrap(~ID, ncol = 4) +
  scale_y_continuous(limits = c(0, 1), breaks = c(0, 0.5, 1)) +
  theme_bw() +
  labs(
    title = "Visualizing 14 Flagged Atypical Participants (T1)",
    subtitle = "Check if they are true flat lines/random guessers or just shallower slopes",
    x = "VOT (ms)",
    y = "Proportion of /t/ Response"
  )

# check performance averaged across all blocks for flagged ids
raw %>%
  filter(trial_type != "catch") %>%
  filter(ID %in% flagged.atypical.cat.ids) %>%
  ggplot(aes(x = VOT, y = resp.num)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", se = FALSE, color = "firebrick") +
  facet_wrap(~ID, ncol = 4) +
  scale_y_continuous(limits = c(0, 1), breaks = c(0, 0.5, 1)) +
  theme_bw() +
  labs(
    title = "Visualizing 14 Flagged Atypical Participants (all blocks)",
    subtitle = "Check if they are true flat lines/random guessers or just shallower slopes",
    x = "VOT (ms)",
    y = "Proportion of /t/ Response"
  )

#-----

exclude_ids <- bind_rows(
  data.exclude.catch.low.acc,
  data.exclude.labelled.low.acc,
  data.exclude.high.na,
  data.exclude.by.subj.rt
) %>% 
  pull(ID) %>% 
  unique()

unique(exclude_ids)
length(unique(exclude_ids))

raw.subj.exclu <- raw %>% 
  filter(!ID %in% exclude_ids)

n_distinct(raw$ID) #N=129 originally
n_distinct(raw.subj.exclu$ID) #N=121 after exclusion

```

```{r trial-level check and exclusion}

#check within-subject RT
raw.subj.exclu.trim.trial <- raw.subj.exclu |> 
   group_by(ID) |> 
   mutate(
     within.subj.rt.log.z = as.numeric(scale(rt.log)),
     exclude.trial.rt = abs(within.subj.rt.log.z) > 3
     ) |>
   ungroup()

table(raw.subj.exclu.trim.trial$exclude.trial.rt) #287/28314=1.01%

df_final <- raw.subj.exclu.trim.trial %>% 
  filter(!exclude.trial.rt)

```

```{r check experiment stats after data exclusion}

#check counterbalanced list assignment
update.sum.list.assign <- df_final %>%
  group_by(list) %>%
  summarise(n_subjects = n_distinct(ID))

#check exposure condition assignment
update.sum.condition.assign <- df_final %>%
  filter(condition!="") %>%
  group_by(condition) %>%
  summarise(n_subjects = n_distinct(ID))

```

#---preliminary data analysis and visualization below---

```{r}

participant.condition <- df_final %>%
  filter(condition != "") %>%
  distinct(ID, condition)

data.plot <- df_final %>%
  filter(trial_type != "catch") %>%
 # left_join(participant.condition, by = "ID") %>%
  mutate(blockID = factor(blockID, levels = c("T1","A","T2","B","T3","C","T4","T5","T6")), ID = factor(ID)) %>%
  group_by(ID, blockID) %>%
  summarise(mean.prop = mean(resp.num), .groups = "drop")

ggplot(data.plot, aes(blockID, mean.prop, colour = ID, group = ID)) +
  geom_line() + 
  geom_point() +
  labs(x = "Block", y = "Proportion /t/")

```

```{r prepare dataset for crticial trials}

df.critical.trial <- df_final %>%
  filter(trial_type != "catch", !is.na(resp.num)) %>%
 # left_join(participant.condition, by = "ID") %>%
  mutate(ID = as.factor(ID)) %>%
  group_by(ID) %>% 
  mutate(
    block_order = match(blockID, unique(blockID)),
    
    exp_stageID = c("Test 1", "Exposure 1", 
                         "Test 2", "Exposure 2", 
                         "Test 3", "Exposure 3", 
                         "Test 4", "Test 5", "Test 6")[block_order],
  ) %>%
  ungroup()  %>%
  mutate(
    exp_stageID = factor(
      exp_stageID, 
      levels = c("Test 1", "Exposure 1", 
                 "Test 2", "Exposure 2", 
                 "Test 3", "Exposure 3", 
                 "Test 4", "Test 5", "Test 6")
    )
  ) %>%
  filter(
    !str_detect(exp_stageID, "Exposure") | 
    (str_detect(exp_stageID, "Exposure") & label_type == "unlabeled")
  ) %>% #for exposure blocks, only unlabeled trials were included for analysis
  mutate(
    minimal_pair = case_when(
      word %in% c("dip", "tip")   ~ "dip-tip",
      word %in% c("din", "tin")   ~ "din-tin",
      word %in% c("dill", "till") ~ "dill-till",
      TRUE ~ word
    )
  )



#separate exposure and test dataset
df_exposure <- df.critical.trial %>% 
  filter(str_detect(trial_type, "exposure")) %>%
  droplevels()

df_test <- df.critical.trial %>% 
  filter(str_detect(trial_type, "test")) %>%
  droplevels()

hist(df_exposure$rt, breaks=50)
hist(df_test$rt, breaks=50)

```
#Parameters and prior setup (see TJ2025 Supplementary material §5.1)
We fit the two psychometric models for exposure and test blocks using the package brms (Bürkner,2017) in R (R Core Team, 2022; RStudio Team, 2020).Predictor coding and priors were identical across both models. To facilitate comparison of effect sizes across predictors, we standardized continuous predictors (VOT) by dividing through twice their standard deviation (Gelman, 2008). We also centered VOT based on the mean of the test blocks. This makes sure that all other effects—e.g., the effects of exposure condition and block—are analyzed at the same VOT across the two separate models. Following previous work from our lab (Hörberg & Jaeger, 2021; Xie, Liu, & Jaeger, 2021) we used weakly regularizing priors to facilitate model convergence. For fixed effect parameters, we used Student priors centered around zero with a scale of 2.5 units (following Gelman et al., 2008) and 3 degrees of freedom, with the exception of VOT which had a wider scale of 15 units because of convergence issues (see §5.5 for elaboration). For the lapse rate we used the brms default logistic prior centered around 0 with a scale of 1 unit. This assumes uniformity for 0 < lapse-rate < 1 while down-weighting the extreme values. For random effect standard deviations, we used a Cauchy prior with location 0 and scale 2, and for random effect correlations, we used an uninformative LKJ-Correlation prior with its only parameter set to 1, describing a uniform prior over correlation matrices (Lewandowski et al., 2009). Four chains with 2000 warm-up samples and 2000 posterior samples each were fit. No divergent transitions after warm-up were observed, and all 1 < 𝑅hat < 1.01.

```{r extract parameters for VOT scaling}

d.mean_sd_scaling <-
  df.critical.trial %>%
  group_by(trial_type) %>%
  #filter(Item.Labeled == FALSE) %>%
  summarise(across(VOT, list(mean = mean, sd = sd)))

VOT.mean_exposure <- (d.mean_sd_scaling %>% pull(VOT_mean))[1]
VOT.sd_exposure <- (d.mean_sd_scaling %>% pull(VOT_sd))[1]
VOT.mean_test <- (d.mean_sd_scaling %>% pull(VOT_mean))[2]
VOT.sd_test <- (d.mean_sd_scaling %>% pull(VOT_sd))[2]
rm(d.mean_sd_scaling)

levels_Condition.Exposure = c("expo-20ms", "expo-10ms", "expo+20ms")

df_exposure <- df_exposure |> 
  mutate(
    VOT_gs = (VOT - VOT.mean_test) / (2 * VOT.sd_exposure)
  )

df_test <- df_test |> 
  mutate(
    VOT_gs = (VOT - VOT.mean_test) / (2 * VOT.sd_test)
  )

```

```{r backward difference coding for exposure condition and block, include=FALSE}

#exposure data
df_exposure$participant.condition <- factor(
  df_exposure$participant.condition, 
  levels = c("expo-20ms", "expo-10ms", "expo+20ms")
)
contrasts(df_exposure$participant.condition) <- contr.sdif(3)

df_exposure$exp_stageID <- factor(
  df_exposure$exp_stageID, 
  levels = c("Exposure 1", "Exposure 2", "Exposure 3")
)
contrasts(df_exposure$exp_stageID) <- contr.sdif(3)

contrasts(df_exposure$participant.condition)
contrasts(df_exposure$exp_stageID)


#test data
df_test$participant.condition <- factor(
  df_test$participant.condition, 
  levels = c("expo-20ms", "expo-10ms", "expo+20ms")
)
contrasts(df_test$participant.condition) <- contr.sdif(3)

df_test$exp_stageID <- factor(
  df_test$exp_stageID, 
  levels = c("Test 1", "Test 2", "Test 3", "Test 4", "Test 5", "Test 6")
)
contrasts(df_test$exp_stageID) <- contr.sdif(6)

contrasts(df_test$participant.condition)
contrasts(df_test$exp_stageID)

```

```{r setup prior, include=FALSE}

my_priors <- c(
  prior(student_t(3, 0, 2.5), class = "b", dpar = "mu2"),
  prior(student_t(3, 0, 15), class = "b", coef = "VOT_gs", dpar = "mu2"),
  prior(cauchy(0, 2), class = "sd", dpar = "mu2"),
  prior(lkj(1), class = "cor"),
  prior(logistic(0, 1), class = "Intercept", dpar = "theta1")
)

```

```{r fit brms model to exposure and test data, include=FALSE}

fit_exposure <- 
  brm(
    bf(resp.num ~ 1,
       mu1 ~ 0 + offset(0),
       mu2 ~ 1 + VOT_gs * participant.condition * exp_stageID + 
       (1 + VOT_gs * exp_stageID | ID) + (1 + VOT_gs * participant.condition * exp_stageID | minimal_pair),
       theta1 ~ 1),
    data = df_exposure,
    prior = my_priors,
    cores = 4,
    chains = 4,
    init = 0,
    iter = 4000,
    warmup = 2000,
    save_pars = save_pars(all = TRUE),
    family = mixture(bernoulli("logit"), bernoulli("logit"), order = F),
    control = list(adapt_delta = 0.99),
    file = "/Users/willchang/Dropbox/Research/UC Irvine/Research/research projects/Xin/NIH accent adaptation/brms-model-results-rds-files/fit-exposure-brms.rds",
    file_refit = "on_change"
  )

summary(fit_exposure)
fit_exposure <- readRDS("/Users/willchang/Dropbox/Research/UC Irvine/Research/research projects/Xin/NIH accent adaptation/brms-model-results-rds-files/fit-exposure-brms.rds")

sink("/Users/willchang/Desktop/git-repo-will-local/Spanish-DL/results/fit-exposure-brms.txt")
print(summary(fit_exposure), digits=3)
sink()

fit_test <- 
  brm(
    bf(resp.num ~ 1,
       mu1 ~ 0 + offset(0),
       mu2 ~ 1 + VOT_gs * participant.condition * exp_stageID + 
       (1 + VOT_gs * exp_stageID | ID) + (1 + VOT_gs * participant.condition * exp_stageID | minimal_pair),
       theta1 ~ 1),
    data = df_test,
    prior = my_priors,
    cores = 4,
    chains = 4,
    init = 0,
    iter = 4000,
    warmup = 2000,
    save_pars = save_pars(all = TRUE),
    family = mixture(bernoulli("logit"), bernoulli("logit"), order = F),
    control = list(adapt_delta = 0.99),
    file = "/Users/willchang/Dropbox/Research/UC Irvine/Research/research projects/Xin/NIH accent adaptation/brms-model-results-rds-files/fit-test-brms.rds",
    file_refit = "on_change"
  )

summary(fit_test)
fit_test <- readRDS("/Users/willchang/Dropbox/Research/UC Irvine/Research/research projects/Xin/NIH accent adaptation/brms-model-results-rds-files/fit-test-brms.rds")

sink("/Users/willchang/Desktop/git-repo-will-local/Spanish-DL/results/fit-test-brms.txt")
print(summary(fit_test), digits=3)
sink()

```

```{r lapse rate for exposure and test blocks, include=FALSE}

#exposure blocks
exposure_lapse_draws <- fit_exposure %>%
  spread_draws(b_theta1_Intercept) %>%
  mutate(lapse_rate = plogis(b_theta1_Intercept))

exposure_lapse_mean_ci <- exposure_lapse_draws %>%
  summarise(
    lapse_median = mean(lapse_rate),
    lapse_lower  = quantile(lapse_rate, 0.025),
    lapse_upper  = quantile(lapse_rate, 0.975)
  )
#mean: 1.81%, 95%-CI: 0.99%-2.79%


#test blocks
test_lapse_draws <- fit_test %>%
  spread_draws(b_theta1_Intercept) %>%
  mutate(lapse_rate = plogis(b_theta1_Intercept))

test_lapse_mean_ci <- test_lapse_draws %>%
  summarise(
    lapse_median = mean(lapse_rate),
    lapse_lower  = quantile(lapse_rate, 0.025),
    lapse_upper  = quantile(lapse_rate, 0.975)
  )
#mean: 3.73%, 95%-CI: 2.60%-4.97%

```

#exposure blocks
```{r extract group-level PSE and slope from brms estimates - exposure blocks}

exposure_pred_grid <- expand_grid(
  participant.condition = unique(df_exposure$participant.condition),
  exp_stageID = unique(df_exposure$exp_stageID),
  VOT_gs = seq(min(df_exposure$VOT_gs), max(df_exposure$VOT_gs), length.out = 100)
)

exposure_pse_draws <- exposure_pred_grid %>%
  add_epred_draws(fit_exposure, re_formula = NA) %>% #default to all draws if ndraws is not specified
  group_by(participant.condition, exp_stageID, .draw) %>%
  slice_min(abs(.epred - 0.5), n = 1) %>%
  mutate(
    PSE_unscaled = (VOT_gs * 2 * VOT.sd_exposure) + VOT.mean_test
  )

saveRDS(exposure_pse_draws, "/Users/willchang/Dropbox/Research/UC Irvine/Research/research projects/Xin/NIH accent adaptation/brms-model-results-rds-files/exposure-pse-draws-predictions.rds")
exposure_pse_draws <- readRDS("/Users/willchang/Dropbox/Research/UC Irvine/Research/research projects/Xin/NIH accent adaptation/brms-model-results-rds-files/exposure-pse-draws-predictions.rds")

d.exposure_PSE_summary_group <- exposure_pse_draws %>%
  group_by(participant.condition, exp_stageID) %>%
  summarise(
    group_PSE_median = median(PSE_unscaled),
    group_PSE_lower  = quantile(PSE_unscaled, 0.025),
    group_PSE_upper  = quantile(PSE_unscaled, 0.975),
    ci_width         = group_PSE_upper - group_PSE_lower,
    .groups = "drop"
  )

write_csv(d.exposure_PSE_summary_group, "/Users/willchang/Desktop/git-repo-will-local/Spanish-DL/results/d.exposure_PSE_summary_group.csv")

```

```{r group-level categorization curves - exposure blocks}

exposure_curve_draws <- exposure_pred_grid %>%
  add_epred_draws(fit_exposure, re_formula = NA) %>%
  mutate(VOT_ms = (VOT_gs * 2 * VOT.sd_exposure) + VOT.mean_test)

saveRDS(exposure_curve_draws, "/Users/willchang/Dropbox/Research/UC Irvine/Research/research projects/Xin/NIH accent adaptation/brms-model-results-rds-files/exposure-curve-draws-predictions.rds")
exposure_curve_draws <- readRDS("/Users/willchang/Dropbox/Research/UC Irvine/Research/research projects/Xin/NIH accent adaptation/brms-model-results-rds-files/exposure-curve-draws-predictions.rds")

exposure_curve_summary_group <- exposure_curve_draws %>%
  mutate(VOT_ms = (VOT_gs * 2 * VOT.sd_exposure) + VOT.mean_test) %>%
  group_by(VOT_ms, participant.condition, exp_stageID) %>%
  summarise(
    pred_prob = median(.epred),
    lower_95  = quantile(.epred, 0.025),
    upper_95  = quantile(.epred, 0.975),
    .groups   = "drop"
  )

ggplot(exposure_curve_summary_group, aes(x = VOT_ms, y = pred_prob, color = participant.condition, fill = participant.condition)) +
  geom_ribbon(aes(ymin = lower_95, ymax = upper_95), alpha = 0.2, color = NA) +
  geom_line(size = 1.2) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray50") +
  facet_wrap(~ exp_stageID) +
  labs(
    x = "VOT (ms)",
    y = "Probability of Response (Logit-fitted)",
    color = "Exposure Condition",
    fill  = "Exposure Condition",
    title = "Psychometric S-Curves across Conditions and Blocks"
  ) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
  theme_bw(base_size = 14) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )

```

```{r extract individual-level PSE and slope from brms estimates - exposure blocks}

id_condition_map <- df_exposure %>% 
  distinct(ID, participant.condition) %>%
  mutate(ID = as.character(ID))

fixef_draws <- fit_exposure %>%
  spread_draws(
    b_mu2_Intercept,
    b_mu2_VOT_gs,
    b_mu2_participant.condition2M1,
    b_mu2_participant.condition3M2,
    b_mu2_exp_stageID2M1,
    b_mu2_exp_stageID3M2,
    `b_mu2_VOT_gs:exp_stageID2M1`,
    `b_mu2_VOT_gs:exp_stageID3M2`
  )

ranef_draws <- fit_exposure %>%
  spread_draws(r_ID__mu2[ID, term]) %>%
  pivot_wider(names_from = term, values_from = r_ID__mu2) %>%
  mutate(ID = as.character(ID))

expoure_draws_ind <- inner_join(fixef_draws, ranef_draws, by = c(".chain", ".iteration", ".draw")) %>%
  left_join(id_condition_map, by = "ID")

exposure_pse_draws_ind_full <- expoure_draws_ind %>%
  mutate(
    cond_offset = case_when(
      participant.condition == "expo-20ms" ~ 0,
      participant.condition == "expo-10ms" ~ b_mu2_participant.condition2M1,
      participant.condition == "expo+20ms" ~ b_mu2_participant.condition2M1 + b_mu2_participant.condition3M2,
      TRUE ~ 0
    ),
    
    rand_slp = if("VOT_gs" %in% names(.)) VOT_gs else 0,
    
    Int_S1 = b_mu2_Intercept + cond_offset + Intercept,
    Int_S2 = Int_S1 + b_mu2_exp_stageID2M1,
    Int_S3 = Int_S2 + b_mu2_exp_stageID3M2,
    
    Slope_S1 = b_mu2_VOT_gs + rand_slp,
    Slope_S2 = Slope_S1 + `b_mu2_VOT_gs:exp_stageID2M1`,
    Slope_S3 = Slope_S2 + `b_mu2_VOT_gs:exp_stageID3M2`
  ) %>%
  
  pivot_longer(
    cols = c(Int_S1, Int_S2, Int_S3),
    names_to = "exp_stageID",
    names_prefix = "Int_S",
    values_to = "indiv_Intercept"
  ) %>%
  mutate(
    indiv_Slope = case_when(
      exp_stageID == "1" ~ Slope_S1,
      exp_stageID == "2" ~ Slope_S2,
      exp_stageID == "3" ~ Slope_S3
    ),
    
    PSE_scaled   = -indiv_Intercept / indiv_Slope,
    PSE_unscaled = (PSE_scaled * 2 * VOT.sd_exposure) + VOT.mean_test
  )

exposure_pse_summary_ind <- exposure_pse_draws_ind_full %>%
  group_by(ID, participant.condition, exp_stageID) %>%
  summarise(
    pse_median = median(PSE_unscaled, na.rm = TRUE),
    pse_lower  = quantile(PSE_unscaled, 0.025, na.rm = TRUE),
    pse_upper  = quantile(PSE_unscaled, 0.975, na.rm = TRUE),
    ci_width   = pse_upper - pse_lower,
    .groups    = "drop"
  )

high_ci_ids <- exposure_pse_summary_ind %>%
  filter(ci_width > 50)

high_ci_ids_list <- high_ci_ids %>% 
  pull(ID) %>%
  unique()

low_ci_ids <- exposure_pse_summary_ind %>%
  filter(ci_width < 25)

low_ci_ids_list <- low_ci_ids %>%
  pull(ID) %>%
  unique()

#--

#individual-level VOT slope
exposure_slope_draws_ind <- exposure_pse_draws_ind_full %>%
  mutate(
    indiv_Slope_ms = indiv_Slope / (2 * VOT.sd_exposure)
  )

exposure_slope_summary_ind <- exposure_slope_draws_ind %>%
  group_by(ID, participant.condition, exp_stageID) %>%
  summarise(
    slope_median_logit = median(indiv_Slope, na.rm = TRUE),
    slope_median_ms    = median(indiv_Slope_ms, na.rm = TRUE),
    slope_lower_ms     = quantile(indiv_Slope_ms, 0.025, na.rm = TRUE),
    slope_upper_ms     = quantile(indiv_Slope_ms, 0.975, na.rm = TRUE),
    .groups            = "drop"
  )

zero_crossing_slope_ids <- exposure_slope_summary_ind %>%
  filter(slope_lower_ms <= 0 & slope_upper_ms >= 0) 

zero_crossing_slope_ids_list <- zero_crossing_slope_ids %>%
  pull(ID) %>%
  unique()

```

```{r plotting curves for bad and good participants - exposure blocks}

exposure_bad_ids_both <- intersect(high_ci_ids_list, zero_crossing_slope_ids_list)
exposure_bad_ids_either <- union(high_ci_ids_list, zero_crossing_slope_ids_list)
exposure_good_ids <- low_ci_ids_list

target_ids <- exposure_bad_ids_either

df_target_raw <- df_exposure %>%
  filter(as.character(ID) %in% target_ids) %>%
  mutate(exp_stageID = as.character(exp_stageID))
  
  #logistic curve: Prob = 1 / (1 + exp(-(Int + Slope * VOT))))
  target_fitted_curves <- exposure_pse_draws_ind_full %>%
    filter(ID %in% target_ids) %>%
    group_by(ID, participant.condition, exp_stageID) %>%
    summarise(
      mid_Int   = median(indiv_Intercept, na.rm = TRUE),
      mid_Slope = median(indiv_Slope, na.rm = TRUE),
      .groups   = "drop"
    ) %>%

    expand_grid(
      VOT_gs = seq(min(df_exposure$VOT_gs, na.rm = TRUE), 
                   max(df_exposure$VOT_gs, na.rm = TRUE), 
                   length.out = 100)
    ) %>%
    mutate(
      VOT_ms = (VOT_gs * 2 * VOT.sd_exposure) + VOT.mean_test,
      prob   = 1 / (1 + exp(-(mid_Int + mid_Slope * VOT_gs))),
      exp_stageID = paste("Exposure", exp_stageID)
    )

  ggplot() +
    #average propotion of /t/ response at each VOT step
    stat_summary_bin(
      data = df_target_raw,
      aes(x = (VOT_gs * 2 * VOT.sd_exposure) + VOT.mean_test, 
          y = resp.num,
          color = exp_stageID),
      fun = mean, geom = "point", size = 2, alpha = 0.6
    ) +
    geom_line(
      data = target_fitted_curves,
      aes(x = VOT_ms, y = prob, color = exp_stageID),
      linewidth = 1
    ) +
    geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray50") +
    facet_wrap(~ ID, scales = "free_x") +
    labs(
      #title = "Atypical Individual Psychometric Curves",
      #subtitle = "Overlaying raw response mean proportions with fitted Bayesian individual curves",
      x = "VOT (ms)",
      y = "Probability of Response",
      color = "Exposure Stage"
    ) +
    theme_bw() +
    theme(legend.position = "bottom")

```

#test blocks
```{r extract group-level PSE and slope from brms estimates - test blocks}

test_pred_grid <- expand_grid(
  participant.condition = unique(df_test$participant.condition),
  exp_stageID = unique(df_test$exp_stageID),
  VOT_gs = seq(min(df_test$VOT_gs), max(df_test$VOT_gs), length.out = 100)
)

test_pse_draws <- test_pred_grid %>%
  add_epred_draws(fit_test, re_formula = NA) %>% #default to all draws if ndraws is not specified
  group_by(participant.condition, exp_stageID, .draw) %>%
  slice_min(abs(.epred - 0.5), n = 1) %>%
  mutate(
    PSE_unscaled = (VOT_gs * 2 * VOT.sd_test) + VOT.mean_test
  )
saveRDS(test_pse_draws, "/Users/willchang/Dropbox/Research/UC Irvine/Research/research projects/Xin/NIH accent adaptation/brms-model-results-rds-files/test-pse-draws-predictions.rds")
test_pse_draws <- readRDS("/Users/willchang/Dropbox/Research/UC Irvine/Research/research projects/Xin/NIH accent adaptation/brms-model-results-rds-files/test-pse-draws-predictions.rds")

d.test_PSE_summary_group <- test_pse_draws %>%
  group_by(participant.condition, exp_stageID) %>%
  summarise(
    group_PSE_median = median(PSE_unscaled),
    group_PSE_lower  = quantile(PSE_unscaled, 0.025),
    group_PSE_upper  = quantile(PSE_unscaled, 0.975),
    ci_width         = group_PSE_upper - group_PSE_lower,
    .groups = "drop"
  )

write_csv(d.test_PSE_summary_group, "/Users/willchang/Desktop/git-repo-will-local/Spanish-DL/results/d.test_PSE_summary_group.csv")

```

```{r group-level categorization curves - test blocks}

test_curve_draws <- test_pred_grid %>%
  add_epred_draws(fit_test, re_formula = NA) %>%
  mutate(VOT_ms = (VOT_gs * 2 * VOT.sd_test) + VOT.mean_test)

saveRDS(test_curve_draws, "/Users/willchang/Dropbox/Research/UC Irvine/Research/research projects/Xin/NIH accent adaptation/brms-model-results-rds-files/test-curve-draws-predictions.rds")
test_curve_draws <- readRDS("/Users/willchang/Dropbox/Research/UC Irvine/Research/research projects/Xin/NIH accent adaptation/brms-model-results-rds-files/test-curve-draws-predictions.rds")

test_curve_summary_group <- test_curve_draws %>%
  mutate(VOT_ms = (VOT_gs * 2 * VOT.sd_test) + VOT.mean_test) %>%
  group_by(VOT_ms, participant.condition, exp_stageID) %>%
  summarise(
    pred_prob = median(.epred),
    lower_95  = quantile(.epred, 0.025),
    upper_95  = quantile(.epred, 0.975),
    .groups   = "drop"
  )

ggplot(test_curve_summary_group, aes(x = VOT_ms, y = pred_prob, color = participant.condition, fill = participant.condition)) +
  geom_ribbon(aes(ymin = lower_95, ymax = upper_95), alpha = 0.2, color = NA) +
  geom_line(size = 1.2) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray50") +
  facet_wrap(~ exp_stageID) +
  labs(
    x = "VOT (ms)",
    y = "Probability of Response (Logit-fitted)",
    color = "Exposure Condition",
    fill  = "Exposure Condition",
    title = "Psychometric S-Curves across Conditions and Blocks"
  ) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
  theme_bw(base_size = 14) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )

```

```{r combine curve plots - A}

all_curve_summary <- bind_rows(exposure_curve_summary_group, test_curve_summary_group)
stages_order_A <- c("Test 1", "Exposure 1", "Test 2", "Exposure 2", "Test 3", "Exposure 3", "Test 4")

all_curve_grid_A <- all_curve_summary %>%
  mutate(exp_stageID = factor(exp_stageID, levels = stages_order_A))

ggplot(all_curve_grid_A %>% filter(!is.na(exp_stageID)),
       aes(x = VOT_ms, y = pred_prob, color = participant.condition, fill = participant.condition)) +
  geom_ribbon(aes(ymin = lower_95, ymax = upper_95), alpha = 0.2, color = NA) +
  geom_line(size = 1.2) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray50") +
  facet_wrap(~ exp_stageID, nrow = 1, scales = "free_x") +
  labs(
    x = "VOT (ms)",
    y = "Probability of Response (Logit-fitted)",
    color = "Exposure Condition",
    fill  = "Exposure Condition",
    title = "Psychometric S-Curves across Conditions and Blocks"
  ) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
  theme_bw(base_size = 14) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )

```

```{r combine plots - B}

stages_order_B <- c("Test 4", "Test 5", "Test 6")

all_curve_grid_B <- all_curve_summary %>%
  mutate(exp_stageID = factor(exp_stageID, levels = stages_order_B))

ggplot(all_curve_grid_B %>% filter(!is.na(exp_stageID)),
       aes(x = VOT_ms, y = pred_prob, color = participant.condition, fill = participant.condition)) +
  geom_ribbon(aes(ymin = lower_95, ymax = upper_95), alpha = 0.2, color = NA) +
  geom_line(size = 1.2) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray50") +
  facet_wrap(~ exp_stageID, nrow = 1, scales = "free_x") +
  labs(
    x = "VOT (ms)",
    y = "Probability of Response (Logit-fitted)",
    color = "Exposure Condition",
    fill  = "Exposure Condition",
    title = "Psychometric S-Curves across Conditions and Blocks"
  ) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
  theme_bw(base_size = 14) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )

```

```{r define expected learning in idealized listeners' based on TJ2025, Figure S6}

baseline_prior <- 44.53 #roughly

baseline_test1 <- d.test_PSE_summary_group %>%
  filter(exp_stageID == "Test 1") %>%
  dplyr::select(participant.condition, PSE_test1 = group_PSE_median) #use as baseline to calculate proportion of adaptation

#exposure
ideal_exposure <- tibble(
  participant.condition = c("expo+20ms", "expo-10ms", "expo-20ms"),
  expected_pse_exposure = c(64.49, 34.35, 24.83)
)

#test
ideal_test <- tibble(
  participant.condition = c("expo+20ms", "expo-10ms", "expo-20ms"),
  expected_pse_test = c(64.59, 34.53, 24.49)
)

```

```{r proportion of adaptation (ideal shift endpoint measured against prior)- exposure and test blocks}

pse_exposure_with_pct.1 <- d.exposure_PSE_summary_group %>%
  left_join(baseline_test1, by = "participant.condition") %>%
  left_join(ideal_exposure, by = "participant.condition") %>%
  mutate(
    pct = (group_PSE_median - PSE_test1) / (expected_pse_exposure - PSE_test1),
    pct_label = sprintf("%.1f%%", pct * 100),
    block_type = "Exposure"
  )

pse_test_with_pct.1 <- d.test_PSE_summary_group %>%
  left_join(baseline_test1, by = "participant.condition") %>%
  left_join(ideal_test, by = "participant.condition") %>%
  mutate(
    pct = (group_PSE_median - PSE_test1) / (expected_pse_test - PSE_test1),
    pct_label = sprintf("%.1f%%", pct * 100),
    block_type = "Test"
  )

pse_combined_with_pct.1 <- bind_rows(pse_exposure_with_pct.1, pse_test_with_pct.1) %>%
  mutate(exp_stageID = factor(exp_stageID, levels = c(
    "Test 1", "Exposure 1", 
    "Test 2", "Exposure 2", 
    "Test 3", "Exposure 3", 
    "Test 4", "Test 5", "Test 6"
  ))) %>% 
  arrange(participant.condition, exp_stageID)

write_csv(pse_combined_with_pct.1, "/Users/willchang/Desktop/git-repo-will-local/Spanish-DL/results/pse_combined_with_pct.csv")

```

```{r proportion plot against ideal listeners (ideal shift endpoint measured against prior)}


# Idealized Learner (based on values in TJ2025-supplementary material-figureS6)
ideal_lines <- data.frame(
  participant.condition = c("expo+20ms", "expo-10ms", "expo-20ms"),
  yintercept = c(64.59, 34.53, 24.49),
  label = c("+21 (100%)\nidealized learner", 
            "-9 (100%)\nidealized learner", 
            "-18 (100%)\nidealized learner")
)


p_fig.idealbyPrior <- ggplot(pse_combined_with_pct.1, aes(x = exp_stageID, y = group_PSE_median, group = participant.condition, color = participant.condition)) +
  
  geom_rect(
    xmin = -Inf, xmax = Inf,
    ymin = 43.53, ymax = 45.53,
    fill = "grey85", color = NA, alpha = 0.5,
    inherit.aes = FALSE
  ) +
  
  geom_hline(
    data = ideal_lines, 
    aes(yintercept = yintercept, color = participant.condition),
    linetype = "dashed", alpha = 0.6, linewidth = 0.6) +
  
  geom_text(
    data = ideal_lines,
    aes(x = Inf, y = yintercept, label = label, color = participant.condition),
    hjust = -0.15,
    vjust = 0.5,
    size = 3.2, 
    fontface = "bold",
    lineheight = 0.85,
    inherit.aes = FALSE,
    show.legend = FALSE
  ) +
  
  geom_errorbar(
    aes(ymin = group_PSE_lower, ymax = group_PSE_upper),
    width = 0.15,
    linewidth = 0.7,
    position = position_dodge(0.2)
  ) +

  geom_line(linewidth = 0.9, position = position_dodge(0.2)) +
  geom_point(size = 2.5, position = position_dodge(0.2)) +
  
  geom_label(
    data = filter(pse_combined_with_pct.1, exp_stageID != "Test 1"),
    aes(label = pct_label),
    vjust = -0.6,
    label.padding = unit(0.15, "lines"),
    label.size = 0.2,
    fill = "white",
    alpha = 0.85,
    show.legend = FALSE,
    position = position_dodge(0.2)
  ) +
  
  scale_color_manual(
    breaks = c("expo-20ms", "expo-10ms", "expo+20ms"),
    values = c("expo-20ms" = "#D32F2F", "expo-10ms" = "#2E7D32", "expo+20ms" = "#1E88E5")
  ) +
  
  labs(
    x = "Block",
    y = "PSE (ms VOT)",
    title = "Perceptual Boundary Shift across Experimental Blocks",
  ) +
  coord_cartesian(clip = "off") + 
  
  theme_bw() +
  theme(
    plot.margin = margin(t = 10, r = 85, b = 10, l = 10, unit = "pt"),
    
    axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1, size = 10, color = "black"),
    axis.text.y = element_text(size = 10, color = "black"),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    
    legend.position = "top",
    legend.title = element_blank(),
    legend.text = element_text(size = 10),
    
    panel.grid.major.x = element_line(color = "grey92"),
    panel.grid.minor = element_blank()
  )

print(p_fig.idealbyPrior)

```





#old LMM analysis
```{r}

# categorization response by participants
ggplot(data=df.critical.trial, aes(x=VOT, y=resp.num, colour = ID, group = ID)) +
  stat_summary(fun = mean, geom = "point", 
               position = position_jitter(width = 0.8, height = 0.02), size=1.5) +
  stat_summary(fun = mean, geom = "line", alpha = 0.5) + 
  facet_wrap(~ exp_stageID) +
  labs(x = "VOT (ms)", y = "Proportion /t/") +
  theme_bw()

```

```{r exposure blocks PSE}

#full glm model
fit_exposure <- glmer(
  resp.num ~ VOT * exp_stageID * participant.condition + 
    (1 + VOT | ID) + (1 | stim),
  data = df_exposure,
  family = binomial,
  nAGQ = 0,
  control = glmerControl(optimizer = "bobyqa"),
  verbose = 1
  )

summary(fit_exposure)


#establish predictions for response proportion by VOT, block, and exposure condition
#exposure_preds <- emmeans(
#  fit_exposure, 
#  ~ VOT | exp_stageID * participant.condition,
#  at = list(VOT = seq(-20, 120, by = 0.1)),
#  type = "response",
#  rg.limit = 40000
#) %>% 
#  as.data.frame()


#derive PSE from the VOT value that corresponds to ~p=0.5
#pse_exposure <- exposure_preds %>%
#  group_by(exp_stageID, participant.condition) %>%
#  filter(abs(prob - 0.5) == min(abs(prob - 0.5))) %>%
#  rename(PSE = VOT) %>%
#  ungroup()


#subj-level PSE
exposure_subj_info <- df_exposure %>% 
  distinct(ID, participant.condition)

exposure_subj <- expand_grid(
  exposure_subj_info,
  exp_stageID = unique(df_exposure$exp_stageID),
  VOT = seq(-20, 125, by = 0.1)
)

exposure_subj$prob <- predict(fit_exposure, newdata = exposure_subj, type = "response", re.form = ~ (1 + VOT | ID))

exposure_subj_pse <- exposure_subj %>%
  group_by(ID, exp_stageID, participant.condition) %>%
  filter(abs(prob - 0.5) == min(abs(prob - 0.5))) %>%
  summarise(PSE_subj = mean(VOT), .groups = "drop")

exposure_pse_group_ci <- exposure_subj_pse %>% #get CI from group mean
  group_by(exp_stageID, participant.condition) %>%
  summarise(
    N = n(),
    PSE = mean(PSE_subj, na.rm = TRUE),
    SD = sd(PSE_subj, na.rm = TRUE),
    SE = SD / sqrt(N),
    PSE_grouplowerCI = PSE - 1.96 * SE,
    PSE_groupupperCI = PSE + 1.96 * SE,
    .groups = "drop"
  )

```

```{r test blocks PSE}

#full glm model
fit_test <- glmer(
  resp.num ~ VOT * exp_stageID * participant.condition + 
    (1 + VOT | ID) + (1 | stim),
  data = df_test,
  family = binomial,
  nAGQ = 0,
  control = glmerControl(optimizer = "bobyqa"),
  verbose = 1
  )

summary(fit_test)

#establish predictions for response proportion by VOT, block, and exposure condition
#test_preds <- emmeans(
#  fit_test, 
#  ~ VOT | exp_stageID * participant.condition,
#  at = list(VOT = seq(-20, 120, by = 0.1)),
#  type = "response",
#  rg.limit = 40000
#) %>% 
#  as.data.frame()


#derive PSE from the VOT value that corresponds to ~p=0.5
#pse_test <- test_preds %>%
#  group_by(exp_stageID, participant.condition) %>%
#  filter(abs(prob - 0.5) == min(abs(prob - 0.5))) %>%
#  rename(PSE = VOT) %>%
#  ungroup()


#subj-level PSE
test_subj_info <- df_test %>% 
  distinct(ID, participant.condition)

test_subj <- expand_grid(
  test_subj_info,
  exp_stageID = unique(df_test$exp_stageID),
  VOT = seq(-5, 70, by = 0.1)
)

test_subj$prob <- predict(fit_test, newdata = test_subj, type = "response", re.form = ~ (1 + VOT | ID))

test_subj_pse <- test_subj %>%
  group_by(ID, exp_stageID, participant.condition) %>%
  filter(abs(prob - 0.5) == min(abs(prob - 0.5))) %>%
  summarise(PSE_subj = mean(VOT), .groups = "drop")

test_pse_group_ci <- test_subj_pse %>% #get CI from group mean
  group_by(exp_stageID, participant.condition) %>%
  summarise(
    N = n(),
    PSE = mean(PSE_subj, na.rm = TRUE),
    SD = sd(PSE_subj, na.rm = TRUE),
    SE = SD / sqrt(N),
    PSE_grouplowerCI = PSE - 1.96 * SE,
    PSE_groupupperCI = PSE + 1.96 * SE,
    .groups = "drop"
  )

```

#visualize model-predicted caetgorization curves

```{r model-predicted categorization curve for exposure blocks}

#establish grid for model prediction (vot x exposure condition x block)
exposure_vots <- seq(min(df_exposure$VOT, na.rm=TRUE), max(df_exposure$VOT, na.rm=TRUE), by = 0.5)
exposure_conditions <- unique(df_exposure$participant.condition)
exposure_stages <- unique(df_exposure$exp_stageID)

exposure_pred_grid <- expand.grid(
  VOT = exposure_vots,
  participant.condition = exposure_conditions,
  exp_stageID = exposure_stages
)

#log-odds prediction
exposure_pred_logodds <- predict(
  fit_exposure, 
  newdata = exposure_pred_grid, 
  re.form = NA, 
  se.fit = TRUE
)

#convert log-odds back to probability
exposure_pred_grid <- exposure_pred_grid %>%
  mutate(
    fit_logodds = exposure_pred_logodds$fit,
    se_logodds = exposure_pred_logodds$se.fit,
    pred_resp = plogis(fit_logodds),
    conf.low = plogis(fit_logodds - 1.96 * se_logodds),
    conf.high = plogis(fit_logodds + 1.96 * se_logodds)
  )


exposure_pred_curves <- ggplot() +
  stat_summary(
    data = df.subject.mean %>% filter(exp_stageID %in% exposure_stages),
    aes(x = VOT, y = resp.num, color = participant.condition),
    fun = mean, 
    geom = "point", 
    alpha = 0.6,
    size = 1.8
  ) +
  geom_ribbon(
    data = exposure_pred_grid,
    aes(x = VOT, ymin = conf.low, ymax = conf.high, fill = participant.condition),
    alpha = 0.15,
    color = NA
  ) +
  geom_line(
    data = exposure_pred_grid,
    aes(x = VOT, y = pred_resp, color = participant.condition),
    linewidth = 1
  ) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey50", alpha = 0.7) +
  scale_color_manual(
    breaks = c("expo-20ms", "expo-10ms", "expo+20ms"),
    values = c("expo-20ms" = "#D32F2F", "expo-10ms" = "#2E7D32", "expo+20ms" = "#1E88E5")
  ) +
  scale_fill_manual(
    breaks = c("expo-20ms", "expo-10ms", "expo+20ms"),
    values = c("expo-20ms" = "#D32F2F", "expo-10ms" = "#2E7D32", "expo+20ms" = "#1E88E5")
  ) +
  facet_wrap(~ exp_stageID) +
  labs(
    x = "VOT (ms)", 
    y = "Proportion /t/",
    title = "Model-Fitted Categorization Curves (Exposure Blocks)"
  ) +
  theme_bw() +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    axis.text = element_text(color = "black"),
    strip.background = element_rect(fill = "grey95")
  )

print(exposure_pred_curves)

```

```{r model-predicted categorization curve for test blocks}

#establish grid for model prediction (vot x exposure condition x block)
test_vots <- seq(min(df_test$VOT, na.rm=TRUE), max(df_test$VOT, na.rm=TRUE), by = 0.5)
test_conditions <- unique(df_test$participant.condition)
test_stages <- unique(df_test$exp_stageID)

test_pred_grid <- expand.grid(
  VOT = test_vots,
  participant.condition = test_conditions,
  exp_stageID = test_stages
)

#log-odds prediction
test_pred_logodds <- predict(
  fit_test, 
  newdata = test_pred_grid, 
  re.form = NA, 
  se.fit = TRUE
)

#convert log-odds back to probability
test_pred_grid <- test_pred_grid %>%
  mutate(
    fit_logodds = test_pred_logodds$fit,
    se_logodds = test_pred_logodds$se.fit,
    pred_resp = plogis(fit_logodds),
    conf.low = plogis(fit_logodds - 1.96 * se_logodds),
    conf.high = plogis(fit_logodds + 1.96 * se_logodds)
  )


test_pred_curves <- ggplot() +
  stat_summary(
    data = df.subject.mean %>% filter(exp_stageID %in% test_stages),
    aes(x = VOT, y = resp.num, color = participant.condition),
    fun = mean, 
    geom = "point", 
    alpha = 0.6,
    size = 1.8
  ) +
  geom_ribbon(
    data = test_pred_grid,
    aes(x = VOT, ymin = conf.low, ymax = conf.high, fill = participant.condition),
    alpha = 0.15,
    color = NA
  ) +
  geom_line(
    data = test_pred_grid,
    aes(x = VOT, y = pred_resp, color = participant.condition),
    linewidth = 1
  ) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey50", alpha = 0.7) +
  scale_color_manual(
    breaks = c("expo-20ms", "expo-10ms", "expo+20ms"),
    values = c("expo-20ms" = "#D32F2F", "expo-10ms" = "#2E7D32", "expo+20ms" = "#1E88E5")
  ) +
  scale_fill_manual(
    breaks = c("expo-20ms", "expo-10ms", "expo+20ms"),
    values = c("expo-20ms" = "#D32F2F", "expo-10ms" = "#2E7D32", "expo+20ms" = "#1E88E5")
  ) +
  facet_wrap(~ exp_stageID) +
  labs(
    x = "VOT (ms)", 
    y = "Proportion /t/",
    title = "Model-Fitted Categorization Curves (Test Blocks)"
  ) +
  theme_bw() +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    axis.text = element_text(color = "black"),
    strip.background = element_rect(fill = "grey95")
  )

print(test_pred_curves)

```

```{r combine plots - A}

all_pred_grid <- bind_rows(exposure_pred_grid, test_pred_grid)
stages_order_A <- c("Test 1", "Exposure 1", "Test 2", "Exposure 2", "Test 3", "Exposure 3", "Test 4")

all_pred_grid_A <- all_pred_grid %>%
  mutate(exp_stageID = factor(exp_stageID, levels = stages_order_A))

df.subject.mean_A <- df.subject.mean %>%
  mutate(exp_stageID = factor(exp_stageID, levels = stages_order_A))

ggplot() +
  geom_ribbon(
    data = all_pred_grid_A %>% filter(!is.na(exp_stageID)),
    aes(x = VOT, ymin = conf.low, ymax = conf.high, fill = participant.condition),
    alpha = 0.15, color = NA
  ) +
  geom_line(
    data = all_pred_grid_A %>% filter(!is.na(exp_stageID)),
    aes(x = VOT, y = pred_resp, color = participant.condition),
    linewidth = 0.8
  ) +
  stat_summary(
    data = df.subject.mean_A %>% filter(!is.na(exp_stageID)),
    aes(x = VOT, y = resp.num, color = participant.condition),
    fun = mean, geom = "point", alpha = 0.6, size = 1.5
  ) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey70") +
  scale_color_manual(
    breaks = c("expo-20ms", "expo-10ms", "expo+20ms"),
    values = c("expo-20ms" = "#D32F2F", "expo-10ms" = "#2E7D32", "expo+20ms" = "#1E88E5")
  ) +
  scale_fill_manual(
    breaks = c("expo-20ms", "expo-10ms", "expo+20ms"),
    values = c("expo-20ms" = "#D32F2F", "expo-10ms" = "#2E7D32", "expo+20ms" = "#1E88E5")
  ) +
  facet_wrap(~ exp_stageID, nrow = 1, scales = "free_x") +
  theme_bw() +
  theme(
    legend.position = "top",
    strip.background = element_blank(),
    strip.text = element_text(face = "bold")
  )

```

```{r combine plots - B}

stages_order_B <- c("Test 4", "Test 5", "Test 6")

all_pred_grid_B <- all_pred_grid %>%
  mutate(exp_stageID = factor(exp_stageID, levels = stages_order_B))

df.subject.mean_B <- df.subject.mean %>%
  mutate(exp_stageID = factor(exp_stageID, levels = stages_order_B))

ggplot() +
  geom_ribbon(
    data = all_pred_grid_B %>% filter(!is.na(exp_stageID)),
    aes(x = VOT, ymin = conf.low, ymax = conf.high, fill = participant.condition),
    alpha = 0.15, color = NA
  ) +
  geom_line(
    data = all_pred_grid_B %>% filter(!is.na(exp_stageID)),
    aes(x = VOT, y = pred_resp, color = participant.condition),
    linewidth = 0.8
  ) +
  stat_summary(
    data = df.subject.mean_B %>% filter(!is.na(exp_stageID)),
    aes(x = VOT, y = resp.num, color = participant.condition),
    fun = mean, geom = "point", alpha = 0.6, size = 1.5
  ) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey70") +
  scale_color_manual(
    breaks = c("expo-20ms", "expo-10ms", "expo+20ms"),
    values = c("expo-20ms" = "#D32F2F", "expo-10ms" = "#2E7D32", "expo+20ms" = "#1E88E5")
  ) +
  scale_fill_manual(
    breaks = c("expo-20ms", "expo-10ms", "expo+20ms"),
    values = c("expo-20ms" = "#D32F2F", "expo-10ms" = "#2E7D32", "expo+20ms" = "#1E88E5")
  ) +
  facet_wrap(~ exp_stageID, nrow = 1, scales = "free_x") +
  theme_bw() +
  theme(
    legend.position = "top",
    strip.background = element_blank(),
    strip.text = element_text(face = "bold")
  )

```

#visualize proportion of adaptation

```{r define expected learning in idealized listeners' based on TJ2025, Figure S6}

baseline_prior <- 44.53 #roughly

baseline_test1 <- test_pse_group_ci %>%
  filter(exp_stageID == "Test 1") %>%
  dplyr::select(participant.condition, PSE_test1 = PSE) #use as baseline to calculate proportion of adaptation

#exposure
ideal_exposure <- tibble(
  participant.condition = c("expo+20ms", "expo-10ms", "expo-20ms"),
  expected_pse_exposure = c(64.49, 34.35, 24.83)
)

#test
ideal_test <- tibble(
  participant.condition = c("expo+20ms", "expo-10ms", "expo-20ms"),
  expected_pse_test = c(64.59, 34.53, 24.49)
)

```

```{r proportion of adaptation (ideal shift endpoint measured against prior)- exposure and test blocks}

pse_exposure_with_pct.1 <- exposure_pse_group_ci %>%
  left_join(baseline_test1, by = "participant.condition") %>%
  left_join(ideal_exposure, by = "participant.condition") %>%
  mutate(
    pct = (PSE - PSE_test1) / (expected_pse_exposure - PSE_test1),
    pct_label = sprintf("%.1f%%", pct * 100),
    block_type = "Exposure"
  )

pse_test_with_pct.1 <- test_pse_group_ci %>%
  left_join(baseline_test1, by = "participant.condition") %>%
  left_join(ideal_test, by = "participant.condition") %>%
  mutate(
    pct = (PSE - PSE_test1) / (expected_pse_test - PSE_test1),
    pct_label = sprintf("%.1f%%", pct * 100),
    block_type = "Test"
  )


pse_combined_with_pct.1 <- bind_rows(pse_exposure_with_pct, pse_test_with_pct) %>%
  mutate(exp_stageID = factor(exp_stageID, levels = c(
    "Test 1", "Exposure 1", 
    "Test 2", "Exposure 2", 
    "Test 3", "Exposure 3", 
    "Test 4", "Test 5", "Test 6"
  ))) %>% 
  arrange(participant.condition, exp_stageID)

```

```{r proportion plot against ideal listeners (ideal shift endpoint measured against prior)}


# Idealized Learner (based on values in TJ2025-supplementary material-figureS6)
ideal_lines <- data.frame(
  participant.condition = c("expo+20ms", "expo-10ms", "expo-20ms"),
  yintercept = c(64.59, 34.53, 24.49),
  label = c("+21 (100%)\nidealized learner", 
            "-9 (100%)\nidealized learner", 
            "-18 (100%)\nidealized learner")
)


p_fig.idealbyPrior <- ggplot(pse_combined_with_pct.1, aes(x = exp_stageID, y = PSE, group = participant.condition, color = participant.condition)) +
  
  geom_rect(
    xmin = -Inf, xmax = Inf,
    ymin = 43.53, ymax = 45.53,
    fill = "grey85", color = NA, alpha = 0.5,
    inherit.aes = FALSE
  ) +
  
  geom_hline(
    data = ideal_lines, 
    aes(yintercept = yintercept, color = participant.condition),
    linetype = "dashed", alpha = 0.6, linewidth = 0.6) +
  
  geom_text(
    data = ideal_lines,
    aes(x = Inf, y = yintercept, label = label, color = participant.condition),
    hjust = -0.15,
    vjust = 0.5,
    size = 3.2, 
    fontface = "bold",
    lineheight = 0.85,
    inherit.aes = FALSE,
    show.legend = FALSE
  ) +
  
  geom_errorbar(
    aes(ymin = PSE_grouplowerCI, ymax = PSE_groupupperCI),
    width = 0.15,
    linewidth = 0.7,
    position = position_dodge(0.2)
  ) +

  geom_line(linewidth = 0.9, position = position_dodge(0.2)) +
  geom_point(size = 2.5, position = position_dodge(0.2)) +
  
  geom_label(
    data = filter(pse_combined_with_pct.1, exp_stageID != "Test 1"),
    aes(label = pct_label),
    vjust = -0.6,
    label.padding = unit(0.15, "lines"),
    label.size = 0.2,
    fill = "white",
    alpha = 0.85,
    show.legend = FALSE,
    position = position_dodge(0.2)
  ) +
  
  scale_color_manual(
    breaks = c("expo-20ms", "expo-10ms", "expo+20ms"),
    values = c("expo-20ms" = "#D32F2F", "expo-10ms" = "#2E7D32", "expo+20ms" = "#1E88E5")
  ) +
  
  labs(
    x = "Block",
    y = "PSE (ms VOT)",
    title = "Perceptual Boundary Shift across Experimental Blocks",
  ) +
  coord_cartesian(clip = "off") + 
  
  theme_bw() +
  theme(
    plot.margin = margin(t = 10, r = 85, b = 10, l = 10, unit = "pt"),
    
    axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1, size = 10, color = "black"),
    axis.text.y = element_text(size = 10, color = "black"),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    
    legend.position = "top",
    legend.title = element_blank(),
    legend.text = element_text(size = 10),
    
    panel.grid.major.x = element_line(color = "grey92"),
    panel.grid.minor = element_blank()
  )

print(p_fig.idealbyPrior)

```

```{r proportion of adaptation (ideal shift endpoint measured against test 1) - exposure and test blocks}

#consider ideal delta rather than fixed endpoints in FigS6
ideal_deltas <- tibble(
  participant.condition = c("expo+20ms", "expo-10ms", "expo-20ms"),
  ideal_delta = c(21, -9, -18)
)

pse_combined_with_pct.2 <- pse_combined_with_pct.1 %>%
  left_join(ideal_deltas, by = "participant.condition") %>%
  mutate(
    ideal_endpoint_from_test1 = PSE_test1 + ideal_delta,
    shift_from_test1 = PSE - PSE_test1,
    pct_by_ideal_endpoint_from_test1 = shift_from_test1 / ideal_delta
  )

```

```{r proportion plot against ideal listeners (ideal shift endpoint measured against test 1)}

pse_combined_with_pct.2 <- pse_combined_with_pct.2 %>%
  mutate(
    pct_label = if_else(
      exp_stageID == "Test 1", 
      "", 
      sprintf("%.1f%%", pct_by_ideal_endpoint_from_test1 * 100)
    )
  )

ideal_lines_dynamic <- pse_combined_with_pct.2 %>%
  distinct(participant.condition, ideal_endpoint_from_test1, ideal_delta) %>%
  mutate(
    yintercept = ideal_endpoint_from_test1,
    label = sprintf("%+d (100%%)\nidealized learner", ideal_delta)
  )


p_fig.idealbytest1 <- ggplot(pse_combined_with_pct.2, aes(x = exp_stageID, y = PSE, group = participant.condition, color = participant.condition)) +
  
  geom_rect(
    xmin = -Inf, xmax = Inf,
    ymin = 43.53, ymax = 45.53,
    fill = "grey85", color = NA, alpha = 0.5,
    inherit.aes = FALSE
  ) +
  
  geom_hline(
    data = ideal_lines_dynamic, 
    aes(yintercept = yintercept, color = participant.condition),
    linetype = "dashed", alpha = 0.6, linewidth = 0.6
  ) +
  
  geom_text(
    data = ideal_lines_dynamic,
    aes(x = Inf, y = yintercept, label = label, color = participant.condition),
    hjust = -0.15,
    vjust = 0.5,
    size = 3.2, 
    fontface = "bold",
    lineheight = 0.85,
    inherit.aes = FALSE,
    show.legend = FALSE
  ) +
  
  geom_errorbar(
    aes(ymin = PSE_grouplowerCI, ymax = PSE_groupupperCI),
    width = 0.15,
    linewidth = 0.7,
    position = position_dodge(0.2)
  ) +
  
  geom_line(linewidth = 0.9, position = position_dodge(0.2)) +
  geom_point(size = 2.5, position = position_dodge(0.2)) +
  
  geom_label(
    data = filter(pse_combined_with_pct.2, exp_stageID != "Test 1"),
    aes(label = pct_label),
    vjust = -0.6,
    label.padding = unit(0.15, "lines"),
    label.size = 0.2,
    fill = "white",
    alpha = 0.85,
    show.legend = FALSE,
    position = position_dodge(0.2)
  ) +
  
  scale_color_manual(
    breaks = c("expo-20ms", "expo-10ms", "expo+20ms"),
    values = c("expo-20ms" = "#D32F2F", "expo-10ms" = "#2E7D32", "expo+20ms" = "#1E88E5")
  ) +
  
  labs(
    x = "Block",
    y = "PSE (ms VOT)",
    title = "Perceptual Boundary Shift across Experimental Blocks (Test 1 Baseline)"
  ) +
  coord_cartesian(clip = "off") + 
  
  theme_bw() +
  theme(
    plot.margin = margin(t = 10, r = 85, b = 10, l = 10, unit = "pt"),
    
    axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1, size = 10, color = "black"),
    axis.text.y = element_text(size = 10, color = "black"),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    
    legend.position = "top",
    legend.title = element_blank(),
    legend.text = element_text(size = 10),
    
    panel.grid.major.x = element_line(color = "grey92"),
    panel.grid.minor = element_blank()
  )

print(p_fig.idealbytest1)

```
