---
title: "norming-spanish"
date: "2026-03-14"
output:
  pdf_document:
    toc: true
---

```{r packages, include = FALSE}
library(dplyr)
library(stringr)
library(tidyverse)
library(lme4)
library(magrittr)
library(ggeffects)
```


## Exp info

- Experiment design 

The participants will sign a consent form, go through a headphone screening test, and then move on to the norming experiment. In this experiment,they  will go through 23 VOT steps x 4 repetitions/blocks x 3 word pairs + 12 catch trials  = 288 trials in total. Each participant hears all VOT steps across all word pairs (only within-subject). On catch trials, participants are instructed to press either left or right. 

The VOT steps are as follows: -100, -50, -20, -10, 0, 5, 10, 15, 20, 25, 30, 35, 40, 45, 50, 55, 60, 65, 70, 80, 90, 100, 120

- Randomization

The order of trials was randomized for each participant. Catch trials were distributed randomly throughout the experiment.

Participants were given the opportunity to take breaks after each block. Participants took about 20 minutes to complete all the trials. For each participant, /d/-initial words were either always displayed on the left side or always displayed on the right side. Across participants, this ordering was counter-balanced. 


## Preprocessing and exclusion

- Exclusion standards:

Participants who committed more than 3 errors out of the 12 catch trials will be excluded (N = 2).

Participants with an average RT more than three standard deviations from the mean of the by-participant means (N = 0) will be excluded.

Trials that are three SDs from the participant's mean RT were excluded from analysis (n=80) will be excluded.


```{r preprocess, warning=FALSE}

raw <- read.csv("../data/data_exp_30subj_all_tasks.csv") %>%
  rename(ID = 'Participant.Private.ID') %>%
  rename(stim = 'Spreadsheet..stim') %>%
  rename(block = 'Spreadsheet..randomized_trial')%>%
  rename(rt = 'Reaction.Time')%>%
  filter(Display == "phonetic categorization") %>%
  filter(Response.Type == "response") %>%
  mutate(
    catch = case_when(
      str_detect(stim, "left")  ~ Spreadsheet..left,
      str_detect(stim, "right") ~ Spreadsheet..right,
    ))%>%
  mutate(catch.corr = if_else(catch == Response, 1, 0))%>%
  mutate(VOT = as.numeric(str_extract(stim, "(?<=_)[^.]+"))) %>%
  mutate(resp.num = ifelse(Response == "d", 0, 1))%>%
  mutate(word = str_extract(stim, "^[^_]+" )) 

raw$ID = factor(raw$ID)
summary(raw)
```



```{r exclusion, warning=FALSE}

#check timedOut trial numbers in the 2AFC task (participants with zero timedOut trial will not be included here)
data.timeout.num <- raw %>%
  filter(Screen == "2AFC", Response.Type == "timedOut") %>%
  group_by(ID) %>%
  summarise(num_timeouts = n())

# check acc on catch trials, exclude two participants
# 12 catch trials in total
data.catch.rate <- raw %>% 
  filter(stim %in% c("pleft_ame.mp3", "pright_ame.mp3")) %>%
  group_by(ID) %>%
  summarise(num_of_correct = sum(catch.corr, na.rm = TRUE))

# participants with < 60% accuracy on catch trials will be removed
data.cal.acc <- raw %>% 
  filter(stim %in% c("pleft_ame.mp3", "pright_ame.mp3")) %>%
  group_by(ID) %>%
  summarise(exclude.catch = sum(catch.corr, na.rm = TRUE) < 10, .groups = "drop")

data.exclude.catch <- data.cal.acc %>% 
  filter(exclude.catch == TRUE)%>% 
  distinct(ID)

# check participant-level rt,  all qualitfied
data.rt.part <- raw %>%
  left_join(data.cal.acc, by = "ID") %>% 
  filter(exclude.catch == FALSE) %>% 
  filter(stim != "pleft_ame.mp3" & stim != "pright_ame.mp3") %>%
  group_by(ID) %>% 
  mutate(rt.log = log10(ifelse(rt <= 0, NA, rt)),
    rt.log.z = as.numeric(scale(rt.log)),
    rt.log.z.mean = mean(rt.log.z, na.rm = T), 
    exclude.by.part.rt = ifelse(abs(scale(rt.log.z.mean)) > 3, TRUE, FALSE))

data.exclude.by.part.rt <- data.rt.part %>%
  filter(exclude.by.part.rt== TRUE)

# check trial-level rt, 80 trials excluded
data.rt.trial <- data.rt.part %>%
  mutate(exclude.by.trial.rt = ifelse(abs(rt.log.z) > 3, TRUE, FALSE))

data.exclude.by.trial.rt <- data.rt.trial %>% 
  filter(exclude.by.trial.rt == TRUE) 


# get clean data
data.preprc <- data.rt.trial %>% 
  filter(exclude.by.trial.rt == FALSE | exclude.by.part.rt == FALSE) %>%
  filter(stim != "pleft_ame.mp3" & stim != "pright_ame.mp3")  #exclude practice trials

#remove one participant with flat categorization curve
data.preprc <- data.preprc %>% 
  filter(ID !="15432624")

```


## Visualization

```{r visualization}

# prepare the data for plot
data.plot <- data.preprc %>%
  select(VOT, resp.num, ID, block, word, stim)


# Plot the raw data for visual inspection
p.raw <- ggplot(data.plot, aes(x = factor(ID), y = resp.num)) +
  stat_summary(fun = mean, geom = "line") +
  stat_summary(fun = mean, geom = "point") +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 1)+
  labs(x = "ID", y = "mean")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p.raw

# plot group-level data
p.group <- ggplot(data.plot, aes(x = VOT, y = resp.num)) +
  stat_summary(fun = mean, geom = "line") +
  stat_summary(fun = mean, geom = "point") +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 1)+
  geom_smooth(method = "glm", method.args = list(family = "binomial"))+
  #geom_smooth()+
  labs(x = "VOT", y = "probability of /t/")

p.group

# group by each continua
p.group + facet_wrap(~ word) 

p.group + facet_wrap(~ block) 

p.group + facet_wrap(~ ID)

#zoom in to participants with less typical categorization curves
p.group %+% (data.plot %>% filter(ID == "15432624"))
p.group %+% (data.plot %>% filter(ID == "15432784"))
p.group %+% (data.plot %>% filter(ID == "15432616"))
p.group %+% (data.plot %>% filter(ID == "15432579"))
p.group %+% (data.plot %>% filter(ID == "15432590"))

#remove one participant with overall response proportion around 0.5 across the whole VOT continuum
#(already implemented above)
#data.preprc <- data.preprc %>% 
#  filter(ID !="15432624")

```


## Analysis - full data

Our goal is to evaluate whether categorization functions for the three minimal pairs differs a lot from each other. And one minimal pair indeed differs. 

```{r logistic regression}
data.preprc %<>%
  mutate(VOT.z = as.numeric(scale(VOT)),
  word = factor(word))

model <- glmer(resp.num ~ VOT.z * word + (1 + VOT.z | ID),
  data = data.preprc,
  family = binomial)

summary(model)
```

```{r group plot by different words - full data}
v_mean <- mean(data.preprc$VOT)
v_sd   <- sd(data.preprc$VOT)

full_vot_ms = seq(-100, 120, by = 0.5)
full_vot_z = (full_vot_ms - v_mean) / v_sd


full_pred_fixed <- ggpredict(model, 
                       terms = c(paste0("VOT.z [", paste(full_vot_z, collapse=","), "]"), "word"))
plot_df_fixed <- as.data.frame(full_pred_fixed)

plot_df_fixed <- plot_df_fixed %>%
  mutate(VOT_ms = (x * v_sd) + v_mean)

ggplot(plot_df_fixed, aes(x = VOT_ms, y = predicted, color = group, fill = group)) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2, color = NA) +
  geom_line(size = 1.2) +
  labs(x = "VOT (ms)", y = "Probability of /t/", title = "Group-level Categorization") +
  theme_minimal()
```
```{r group-level categorization curve - full data}
model_with_word_randintercept <- glmer(
  resp.num ~ VOT.z + (1 + VOT.z | ID) + (1 | word), 
  data = data.preprc, 
  family = binomial
)

allword_full_pred_fixed <- ggpredict(model_with_word_randintercept, 
                       terms = c(paste0("VOT.z [", paste(full_vot_z, collapse=","), "]"), "word"))

allword_plot_df_fixed <- as.data.frame(allword_full_pred_fixed) %>%
  group_by(x) %>%
  summarize(
    predicted = mean(predicted),
    conf.low = min(conf.low),
    conf.high = max(conf.high)
  ) %>%
  mutate(VOT_ms = (x * v_sd) + v_mean)

ggplot(allword_plot_df_fixed, aes(x = VOT_ms, y = predicted)) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), 
              fill = "grey80", alpha = 0.4) +
  geom_line(size = 1.2, color = "black") +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "red", alpha = 0.5) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(
    x = "VOT (ms)", 
    y = "Probability of /t/ Response",
    title = "Overall Categorization Curve (Averaged across words)"
  ) +
  theme_minimal()

ggplot(data.preprc, aes(x = VOT, y = as.numeric(as.character(resp.num)))) +
  geom_jitter(height = 0.05, alpha = 0.1) + 
  geom_smooth(method = "glm", method.args = list(family = "binomial"), color = "red") +
  labs(title = "Raw Data Overall Trend", y = "Probability of /t/")

```

```{r group plot overlaid with individual curves - full data}

pred_random <- ggpredict(model, terms = c(paste0("VOT.z [", paste(full_vot_z, collapse=","), "]"), "word", "ID"), type = "random")
                         
plot_df_random <- as.data.frame(pred_random) %>%
  mutate(VOT_ms = (x * v_sd) + v_mean)

ggplot() +
  geom_line(data = plot_df_fixed, 
            aes(x = VOT_ms, 
                y = predicted, 
                color = group), 
            size = 1.2) +
  
  geom_line(data = plot_df_random, 
            aes(x = VOT_ms, 
                y = predicted, 
                group = interaction(facet, group)), 
            color = "grey", alpha = 0.3, size = 0.5) +
  
  #coord_cartesian(xlim = c(-20, 100)) + 
  #scale_x_continuous(breaks = seq(-20, 100, 20)) +
  #scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(x = "VOT (ms)", 
       y = "Probability of /t/ Response") +
  facet_wrap(~group)+
  theme_minimal() +
  theme(legend.position = "bottom")

```

```{r VOT boundary - full data}
m <- mean(data.preprc$VOT)
sd <- sd(data.preprc$VOT)

coef <- fixef(model)

calc_VOT_boundary <- function(intercept, slope, m, sd) {
  z_boundary <- -intercept / slope
  VOT_boundary <- (z_boundary * sd) + m
  return(VOT_boundary)
}

# VOT boundaries
# dill (Baseline)
boundary_dill <- calc_VOT_boundary(coef["(Intercept)"], coef["VOT.z"], m, sd)

# din (Intercept + worddin, Slope + VOT.z:worddin)
boundary_din  <- calc_VOT_boundary(coef["(Intercept)"] + coef["worddin"], 
                                  coef["VOT.z"] + coef["VOT.z:worddin"], 
                                  m, sd)

# dip (Intercept + worddip, Slope + VOT.z:worddip)
boundary_dip  <- calc_VOT_boundary(coef["(Intercept)"] + coef["worddip"], 
                                  coef["VOT.z"] + coef["VOT.z:worddip"], 
                                  m, sd)

cat("--- VOT Boundaries (ms) from full dataset---\n")
cat("dill: ", round(boundary_dill, 2), "ms\n")
cat("din:  ", round(boundary_din, 2), "ms\n")
cat("dip:  ", round(boundary_dip, 2), "ms\n")
cat("-----------------------------------\n")
```


## Analysis - subset data (Exclude VOT steps -100, -50, and 120)
```{r plotting - subset data}
data.preprc_subset <- data.preprc %>% 
  filter(!(VOT %in% c(-100, -50, 120)))

# prepare the data for plot
data.plot_subset <- data.preprc_subset %>%
  select(VOT, resp.num, ID, block, word, stim)


# Plot the raw data for visual inspection
p.raw_subset <- ggplot(data.plot_subset, aes(x = factor(ID), y = resp.num)) +
  stat_summary(fun = mean, geom = "line") +
  stat_summary(fun = mean, geom = "point") +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 1)+
  labs(x = "ID", y = "mean")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p.raw_subset

# plot group-level data
p.group_subset <- ggplot(data.plot_subset, aes(x = VOT, y = resp.num)) +
  stat_summary(fun = mean, geom = "line") +
  stat_summary(fun = mean, geom = "point") +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 1)+
 geom_smooth(method = "glm", method.args = list(family = "binomial"))+
  #geom_smooth()+
  labs(x = "VOT", y = "probability of /t/")

p.group_subset

# group by each continua
p.group_subset + facet_wrap(~ word) 

p.group_subset + facet_wrap(~ block) 

p.group_subset + facet_wrap(~ ID)
```

```{r logistic regression - subset data}
data.preprc_subset %<>%
  mutate(VOT.z = as.numeric(scale(VOT)),
  word = factor(word))

model_subset <- glmer(resp.num ~ VOT.z * word + (1 + VOT.z | ID),
  data = data.preprc_subset,
  family = binomial)

summary(model_subset)
```

```{r group plot by different words - subset data}
v_mean2 <- mean(data.preprc_subset$VOT)
v_sd2   <- sd(data.preprc_subset$VOT)

subset_vot_ms = seq(-20, 100, by = 0.5)
subset_vot_z = (subset_vot_ms - v_mean2) / v_sd2

subset_pred_fixed <- ggpredict(model_subset, 
                       terms = c(paste0("VOT.z [", paste(subset_vot_z, collapse=","), "]"), "word"))
subset_plot_df_fixed <- as.data.frame(subset_pred_fixed)

subset_plot_df_fixed <- subset_plot_df_fixed %>%
  mutate(VOT_ms = (x * v_sd2) + v_mean2)

ggplot(subset_plot_df_fixed, aes(x = VOT_ms, y = predicted, color = group, fill = group)) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2, color = NA) +
  geom_line(size = 1.2) +
  labs(x = "VOT (ms)", y = "Probability of /t/", title = "Group-level Categorization") +
  theme_minimal()
```

```{r group plot overlaid with individual curves - subset data}

subset_pred_random <- ggpredict(model_subset, terms = c(paste0("VOT.z [", paste(subset_vot_z, collapse=","), "]"), "word", "ID"), type = "random")
                         
subset_plot_df_random <- as.data.frame(subset_pred_random) %>%
  mutate(VOT_ms = (x * v_sd2) + v_mean2)

ggplot() +
  geom_line(data = subset_plot_df_fixed, 
            aes(x = VOT_ms, 
                y = predicted, 
                color = group), 
            size = 1.2) +
  
  geom_line(data = subset_plot_df_random, 
            aes(x = VOT_ms, 
                y = predicted, 
                group = interaction(facet, group)), 
            color = "grey", alpha = 0.3, size = 0.5) +
  
  #coord_cartesian(xlim = c(-20, 100)) + 
  #scale_x_continuous(breaks = seq(-20, 100, 20)) +
  #scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(x = "VOT (ms)", 
       y = "Probability of /t/ Response") +
  facet_wrap(~group)+
  theme_minimal() +
  theme(legend.position = "bottom")

```

```{r VOT boundary - subset data}
m_subset <- mean(data.preprc_subset$VOT)
sd_subset <- sd(data.preprc_subset$VOT)

coef_subset <- fixef(model_subset)

calc_VOT_boundary <- function(intercept, slope, m, sd) {
  z_boundary <- -intercept / slope
  VOT_boundary <- (z_boundary * sd) + m
  return(VOT_boundary)
}

# VOT boundaries
# dill (Baseline)
subset_boundary_dill <- calc_VOT_boundary(coef_subset["(Intercept)"], coef_subset["VOT.z"], m_subset, sd_subset)

# din (Intercept + worddin, Slope + VOT.z:worddin)
subset_boundary_din  <- calc_VOT_boundary(coef_subset["(Intercept)"] + coef_subset["worddin"], 
                                  coef_subset["VOT.z"] + coef_subset["VOT.z:worddin"], 
                                  m_subset, sd_subset)

# dip (Intercept + worddip, Slope + VOT.z:worddip)
subset_boundary_dip  <- calc_VOT_boundary(coef_subset["(Intercept)"] + coef_subset["worddip"], 
                                  coef_subset["VOT.z"] + coef_subset["VOT.z:worddip"], 
                                  m_subset, sd_subset)

cat("--- VOT Boundaries (ms) from subset dataset---\n")
cat("dill: ", round(subset_boundary_dill, 2), "ms\n")
cat("din:  ", round(subset_boundary_din, 2), "ms\n")
cat("dip:  ", round(subset_boundary_dip, 2), "ms\n")
cat("-----------------------------------\n")
```



